\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This manual is a record of what the \texttt{apfs-fastindex} project has
learned about the Apple File System on disk and about the specific
problem of enumerating a single volume quickly and correctly. It is not
a guide to the repository. The repository is a working environment;
this manual is the textbook the working environment built up.

The reader the manual is written for is a careful engineer who wants to
read APFS metadata directly. We assume familiarity with copy-on-write
filesystems and with B-trees, but not with APFS specifically. Every
claim that is not common knowledge is sourced or labelled.

\section*{Claim Discipline}

Three labels appear throughout the text.

\begin{description}
  \item[Spec.] Backed by Apple's public reference\footnote{\emph{Apple
    File System Reference}, retrieved 2026-05-13,
    \url{https://developer.apple.com/support/downloads/Apple-File-System-Reference.pdf}.}
    or by another public, durable specification.
  \item[Observation.] Backed by a controlled experiment in this project
    or by behavior that converges across multiple independent
    open-source readers
    (\texttt{linux-apfs-rw}, \texttt{apfs-fuse}, \texttt{dissect.apfs},
    \texttt{go-apfs}, \texttt{libfsapfs}, the Sleuth Kit's APFS
    module).
  \item[Hypothesis.] Plausible, useful for design, not yet proven.
    Anything labelled Hypothesis does not justify a product claim.
\end{description}

Where Spec and Observation disagree, Observation wins. Apple's reference
is one citation among many; converged code is often more decisive about
what bytes actually mean. Where the disagreement matters, the text says
so and names the smallest probe that would settle it.

\section*{Oracle Plurality}

There is no single oracle for an APFS indexer. Each feature has its
own:

\begin{itemize}
  \item Namespace: POSIX traversal of the same volume at the same
    selected state.
  \item Logical size: \texttt{st\_size} per inode (or symlink-target
    byte length for symlinks).
  \item FS-record body: cross-tool field equality between independent
    decoders against the same raw container.
  \item Allocated size: \texttt{st\_blocks * 512}, for cases that have
    been explicitly validated.
  \item Incremental correctness: a fresh full reparse of the same
    selected state.
  \item Boot-root semantics: the user-visible macOS namespace, and only
    when the product mode explicitly targets it.
\end{itemize}

Every result in this manual names the oracle that verified it.

\section*{Fail Closed By Default}

The parser refuses to guess. Outside the tested allowlist, it stops
with a typed error and the caller is responsible for choosing a
fallback. Plausible but unproven is treated as "do not emit." The
discipline is not paranoia: it is the cheapest way to keep silently
wrong rows out of aggregates and out of the user interface.

\section*{What Is And Is Not Here}

The manual presents findings about APFS as a textbook would: each
chapter takes one face of the system, states what is known, gives the
evidence, and notes what remains open. Where the project's
implementation makes a non-obvious choice, the choice is explained from
first principles rather than from code organisation.

What is not here: a tour of the repository, a list of files, a list of
commits, a list of experiments treated as work units rather than as
evidence. Those live in the repository itself; the manual cites them
only where citation supports a claim.

\section*{A Note On The Voice}

We have tried to write as though the reader will need to defend each
claim to a sceptic. That style is at odds with the way reverse
engineering is usually written up, but the project has been disciplined
about evidence, and the manual should reflect that. A sentence that
cannot be defended should not have been written.
