\chapter{Research Method}

\section{Artifact Types}

The project uses three durable research artifact types, defined in
\path{docs/research/000-research-index.md}.

\begin{description}
  \item[RL] Research logs. These are living synthesis documents organized by
  question and decision impact.
  \item[SR] Source reviews. These summarize external public documentation,
  open-source implementation behavior, reverse-engineering notes, and support
  limits.
  \item[EX] Experiments. These are controlled probes with scripts, generated
  artifacts, expected outcomes, observed outcomes, and a distilled conclusion.
\end{description}

Raw command output belongs in experiment artifacts. Durable conclusions belong
in the experiment README and then in the affected research logs.

\section{Staged Gates}

The staged gates are:

\begin{description}
  \item[Gate A] Minimally correct v1 parser: checkpoint selection, object
  resolution, FS tree records, namespace, logical size, and validation.
  \item[Gate B] Support boundary: runtime source classes, encryption, feature
  gates, and fallback.
  \item[Gate C] Safe incremental scanning: identity, subtree reuse, persistence,
  and invalidation.
  \item[Gate D] Product semantics and optimization: snapshots, firmlinks,
  volume groups, and performance model.
\end{description}

This ordering prevents performance work from outrunning correctness and prevents
product claims from outrunning support evidence.

\section{Claim Discipline}

Every durable claim should be tagged mentally as one of:

\begin{itemize}
  \item \textbf{Spec}: backed by Apple/public documentation.
  \item \textbf{Observation}: backed by repo experiments or converging external
  implementations.
  \item \textbf{Hypothesis}: plausible, useful, but not yet proven.
\end{itemize}

The manual should preserve that discipline. It is acceptable for a chapter to
say "not yet known"; that is better than encoding a guess as architecture.

\section{Source Review Cascade}

External evidence is consolidated in \path{docs/research/sources/SR-*.md}. The
current set covers:

\begin{itemize}
  \item \texttt{SR-001} v1 support boundary
  \item \texttt{SR-002} checkpoint, OMAP, and root-discovery contract
  \item \texttt{SR-003} FS record taxonomy
  \item \texttt{SR-004} runtime read path and encryption boundary
  \item \texttt{SR-005} checkpoint validation details for the Rust scanner
  \item \texttt{SR-006} OMAP lookup semantics and failure cases
  \item \texttt{SR-007} object-header fail-closed validation
  \item \texttt{SR-008} FS record layout for native v1 parsing
  \item \texttt{SR-009} compression and logical-size precedence
  \item \texttt{SR-010} snapshots, volume groups, and firmlinks
  \item \texttt{SR-011} encryption and runtime read-path boundary
  \item \texttt{SR-012} format drift and feature-bit allowlisting
  \item \texttt{SR-013} checkpoint-map integrity and ephemeral-object validation
  \item \texttt{SR-014} native FS-record body contract
  \item \texttt{SR-015} xfield layout and alignment
  \item \texttt{SR-016} record-body fail-closed boundary
  \item \texttt{SR-017} logical-size source precedence
  \item \texttt{SR-018} name normalization and case behavior
\end{itemize}

Every \texttt{SR-*} note opens with a bottom line, separates Spec from
Observation from Hypothesis, and ends with a Decision impact line that names
the affected \texttt{RL-*} log and the next exact step. Conflicts between
Apple's PDF and converged open-source implementations carry an explicit
\textbf{Conflict Note} so the smallest decisive probe is identifiable from the
review alone.

\section{Pinned Citation Style}

External citations should pin both the URL and either a commit SHA or
retrieval date. \texttt{SR-015} is the current model: the source list cites
Apple's reference PDF together with permalinks at fixed commits for
\texttt{linux-apfs-rw}, \texttt{apfs-fuse}, \texttt{dissect.apfs},
\texttt{go-apfs}, \texttt{libfsapfs}, and the Sleuth Kit's APFS module. A
claim that three or more independent implementations agree on is treated as
stronger evidence than the PDF on its own.
