\chapter{The Container}

Block zero is the locator. It is not the authoritative container
superblock. Treating block zero as the source of truth is the first
mistake an APFS reader can make; the second is reading the highest
checkpoint without validating it; the third is reading the latest
checkpoint while the volume is changing. This chapter is concerned
with avoiding all three.

\section{Block Zero As Locator}

APFS is copy-on-write at the container level. Each transaction writes a
new container superblock (\texttt{nx\_superblock\_t}, henceforth NXSB)
into the checkpoint descriptor area. Block zero is the stable known
location at which a reader can find the descriptor area. It is updated
lazily; the authoritative live NXSB is one of the copies in the
descriptor ring.

The pieces of block zero a reader needs:

\begin{itemize}
  \item \texttt{block\_size} at offset \texttt{0x24}: validated to be a
    power of two within \texttt{[4096, 65536]}.
  \item \texttt{nx\_xp\_desc\_blocks} at offset \texttt{0x68}: the
    length of the descriptor area in blocks.
  \item \texttt{nx\_xp\_desc\_base} at offset \texttt{0x70}: the first
    block of the descriptor area. The high bit signals a
    \emph{non-contiguous} layout, which v1 does not support and which
    triggers a typed hard stop. A non-contiguous descriptor area
    would require a separate checkpoint-mapping-tree implementation.
\end{itemize}

Block zero is itself an APFS object and must validate as one: NXSB
magic at offset \texttt{0x20}, object type \texttt{NX\_SUPERBLOCK},
object id equal to the constant NX-superblock OID (1), and a valid
Fletcher-64 checksum. A block zero that does not satisfy these is not
an APFS container.

\textbf{Observation.} The \texttt{block\_size} and descriptor base in
block zero must agree with the selected NXSB's view of the same
fields. A discrepancy means block zero is stale relative to a
descriptor write that did not (yet) propagate. The parser rejects
the source rather than guessing which view is correct.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  every node/.style={font=\small},
  block/.style={draw, minimum width=1.05cm, minimum height=0.8cm,
                align=center, font=\footnotesize},
  locator/.style={block, fill=gray!12},
  nxsb/.style={block, fill=blue!10},
  nxsbsel/.style={block, fill=blue!30, very thick},
  map/.style={block, fill=orange!18},
  emptyc/.style={block, fill=white, dashed},
  arrow/.style={-Stealth, thick},
]
  \node[locator, minimum width=1.6cm, minimum height=1.5cm] (b0)
    {block 0\\\scriptsize locator};

  \node[nxsb, right=1.8cm of b0, yshift=0.2cm] (n1) {NXSB\\\scriptsize xid 18};
  \node[nxsb, right=0.05cm of n1] (n2) {NXSB\\\scriptsize xid 19};
  \node[map,  right=0.05cm of n2] (m1) {map};
  \node[nxsb, right=0.05cm of m1] (n3) {NXSB\\\scriptsize xid 20};
  \node[nxsbsel, right=0.05cm of n3] (n4) {NXSB\\\scriptsize xid \textbf{21}};
  \node[map,  right=0.05cm of n4] (m2) {map\\\scriptsize last};
  \node[emptyc, right=0.05cm of m2] (e1) {};
  \node[emptyc, right=0.05cm of e1] (e2) {};

  \node[fit=(n1)(e2), draw=none, label={[font=\footnotesize, yshift=-0.05cm]
        above:checkpoint descriptor area}] (ring) {};

  \draw[arrow] (b0.east) -- ++(0.4,0) |- (ring.west)
    node[midway, above, font=\scriptsize] {desc base};

  \draw[arrow, dashed]
    (n4.south) ++(0,-0.1) -- ++(0,-0.7)
    node[below, font=\scriptsize, align=center]
      {selected NXSB:\\highest checksum-valid xid};
\end{tikzpicture}
\caption{Block zero locates the descriptor ring; the selected NXSB is
  the highest checksum-valid candidate in the ring. Several earlier
  NXSBs may remain valid; \texttt{map} blocks carry checkpoint-map
  contents; trailing slots may be unwritten between commits.}
\end{figure}

\section{Selecting The Checkpoint}

The descriptor area is a contiguous run of \texttt{nx\_xp\_desc\_blocks}
blocks. Some are checkpoint maps (chapter~\ref{ssec:checkpoint-map});
some are NXSB candidates. The selection rule:

\begin{enumerate}
  \item Iterate every block in the descriptor range.
  \item For each, check the NXSB magic at offset \texttt{0x20}.
  \item If the magic matches, validate object type
    (\texttt{NX\_SUPERBLOCK}) and Fletcher-64 checksum.
  \item Record the candidate.
  \item After the scan, pick the candidate with the highest XID.
\end{enumerate}

The highest valid XID is the selected NXSB. Its XID becomes the
\emph{scan XID}, and every subsequent object resolution for the run is
performed at this XID.

Blocks that carry NXSB magic but fail object-type or checksum
validation are recorded as skipped descriptors with the reason. They
are not silently ignored; a future support broadening must justify
why they were skipped.

\textbf{Observation.} Three or four NXSB candidates is typical for a
recently written container. The first execution of EX-15 against the
EX-14 fixture saw four candidates with XIDs 18--21; all four passed
strict validation. Multiple candidates is not a sign of corruption.
Zero candidates is.

\subsection{Why The Highest Valid XID}

The candidates are not chronologically ordered in the descriptor area;
they are written into a ring. The XID is the only monotonic identity
for transaction order. The highest checksum-valid XID is the most
recent transaction the writer completed. A higher XID with an invalid
checksum is a transaction the writer started but did not finish; v1
treats those as failed and ignores them, with the result that
selection is always self-consistent.

The rule has a corollary: a parser that picks a lower XID when the
highest is valid has chosen a stale state for no reason and will
disagree with anything else that picks the latest.

\subsection{Live State Is Not Pinned}

\textbf{Observation.} On a mounted APFS volume that other processes
are using, the highest valid XID moves under the reader. Two reads
of "latest" may return two different transactions. Mixing objects
from different transactions produces an incoherent namespace.
EX-01's churn probe demonstrated this on a mounted lab image.

The v1 fix is structural rather than algorithmic: raw mode only runs
on sources where state can be pinned. Detaching an image fixes the
visible state. So does an explicit caller commitment that a device's
checkpoint will not advance. Live mounted volumes are
readable-but-not-supported; chapter~11 develops the vocabulary.

\section{The Checkpoint Map Ring}
\label{ssec:checkpoint-map}

A selected NXSB references a chain of checkpoint maps
(\texttt{checkpoint\_map\_phys\_t}) in the descriptor area. The map
chain carries the ephemeral object identifier table for the selected
checkpoint: the OIDs and physical addresses of objects that live in
memory between transactions and are written into the checkpoint data
area at commit. The reader needs the map to resolve those ephemerals.

The walk rule:

\begin{enumerate}
  \item Start at \texttt{nx\_xp\_desc\_index} in the selected NXSB.
  \item Read the block; validate as an APFS object with type
    \texttt{OBJECT\_TYPE\_CHECKPOINT\_MAP}.
  \item Record the (\texttt{oid}, \texttt{paddr}) mappings.
  \item Advance through subsequent map blocks in the descriptor area.
  \item Stop when a block carries the \texttt{CHECKPOINT\_MAP\_LAST}
    flag.
\end{enumerate}

A chain that runs out of length before the last flag is malformed.
v1 fails closed.

\textbf{Empirical correction.} The Apple reference describes
ephemeral objects in a way that reads naturally as "the checkpoint
map block itself is ephemeral." It is not. Validation against a
real container demonstrates that \texttt{OBJECT\_TYPE\_CHECKPOINT\_MAP}
blocks carry the \texttt{OBJ\_PHYSICAL} storage flag, not
\texttt{OBJ\_EPHEMERAL}. The validator must expect
\texttt{Physical} storage for these blocks. This correction was
landed during EX-11 as a one-line constant change with a synthetic
malformed-checkpoint-map regression test.

\section{Roots Reachable From The Selected Checkpoint}

The selected NXSB names five things the v1 reader needs:

\begin{enumerate}
  \item \texttt{nx\_omap\_oid}: the OID of the container OMAP.
  \item \texttt{nx\_fs\_oid[]}: an array of volume superblock OIDs.
  \item Feature, incompatible-feature, and read-only-compatible
    feature masks.
  \item Reaper and space-manager OIDs (not used by v1; their presence
    is required, their content is not).
  \item Container UUID (used for source identity in caches and
    diagnostics).
\end{enumerate}

The incompatible-feature mask is the gate for format drift
(chapter~11). Any bit the reader does not recognise stops the run.
Better to refuse a container with a new feature bit than to
silently emit a row that depends on the bit's meaning.

The container OMAP and the volume superblocks both require object
resolution, and that is the next chapter's subject. The point to fix
here is that block zero, the descriptor area, and the checkpoint map
together establish a single coherent transaction context. Everything
that follows runs inside that context. The context is not negotiable.
