\chapter{V1 Raw Pipeline}

\section{Pipeline}

The narrow raw parser pipeline is:

\begin{enumerate}
  \item source gate
  \item checkpoint pinning
  \item container root discovery
  \item volume root discovery
  \item FS record walk
  \item namespace entry construction
  \item logical-size aggregate construction
  \item oracle comparison
\end{enumerate}

The implementation-facing contract is
\path{docs/research/contracts/narrow-v1-parser-contract.md}. The first prototype
plan is \path{docs/research/plans/first-raw-parser-prototype-plan.md}.

\section{Checkpoint And Scan State}

The parser must choose one coherent state and keep all object resolution within
that state. \path{docs/research/RL-01-checkpoint-selection-and-consistency.md}
tracks the open checkpoint questions. \path{docs/research/sources/SR-002-checkpoint-omap-root-contract.md}
is the current source review for checkpoint, OMAP, and root discovery.

\section{OMAP And Object Resolution}

Object lookup is not a global \texttt{oid -> paddr} map. Resolution needs the
owning OMAP context, logical object id, and selected scan state. The active logs
are \path{docs/research/RL-02-omap-and-object-resolution.md} and
\path{docs/research/RL-04-node-identity-cache-keys-and-oid-reuse.md}.

\section{FS Records}

The v1 record surface is intentionally narrow:

\begin{itemize}
  \item directory records for membership and names
  \item inode records for identity and metadata
  \item dstream or equivalent fields for logical size
  \item symlink xattrs for symlink target fidelity
  \item hard-link records where present
\end{itemize}

The current source-backed taxonomy is
\path{docs/research/sources/SR-003-fs-record-taxonomy.md}; the rolling synthesis
is \path{docs/research/RL-03-fs-tree-topology-and-required-records.md}.

\subsection{Body And Xfield Decoding Rules}

FS-record body decoding adds three further source-backed contracts:

\begin{itemize}
  \item \path{docs/research/sources/SR-008-fs-record-layout-v1.md} fixes the
  on-disk layout of \texttt{j\_drec\_*}, \texttt{j\_inode\_*},
  \texttt{j\_dstream\_*}, \texttt{j\_xattr\_*}, and \texttt{j\_sibling\_*}.
  \item \path{docs/research/sources/SR-014-native-fs-record-body-contract.md}
  ties body decoding to the validated checkpoint/OMAP/root context.
  \item \path{docs/research/sources/SR-015-xfield-layout-and-alignment.md}
  pins one deterministic xfield cursor: the value area starts immediately
  after the metadata table and each value advances by
  \texttt{round\_up(x\_size, 8)}; \texttt{xf\_used\_data} is the padded
  value-area byte count.
  \item \path{docs/research/sources/SR-016-record-body-fail-closed-boundary.md}
  enumerates the malformed-body hard stops (short fixed bodies, malformed
  names, duplicate or out-of-bounds xfields, invalid xattr forms,
  \texttt{drec}/\texttt{inode} type mismatches, missing sibling mappings).
\end{itemize}

\section{Namespace And Logical Size}

Namespace output must preserve path identity and file identity separately.
Logical size is the only v1 size metric. Physical, shared, exclusive, and
snapshot-retained accounting are separate modes.

The logical-size source precedence is fixed by
\path{docs/research/sources/SR-017-logical-size-source-precedence.md}: ordinary,
sparse, cloned, hard-linked, symlink, and compressed cases each name the
candidate fields (\texttt{j\_dstream\_t.size}, \texttt{INO\_EXT\_TYPE\_DSTREAM},
\texttt{INO\_EXT\_TYPE\_SPARSE\_BYTES}, decmpfs xattr) and the fail-closed
condition when they disagree.

Name normalization, case behavior, and APFS hash collisions follow
\path{docs/research/sources/SR-018-name-normalization-and-case.md}. Until the
SR-018 fixture lands, stored UTF-8 names are emitted verbatim and lookup-by-name
is not claimed.

Related logs:

\begin{itemize}
  \item \path{docs/research/RL-06-namespace-reconstruction.md}
  \item \path{docs/research/RL-07-size-and-space-accounting.md}
\end{itemize}

\section{Current Open Blocker}

\path{docs/research/experiments/EX-14-xfield-layout-variant/README.md} returned
\texttt{oracle\_inconclusive} on a denser EX-13-style fixture: the Rust context
provider found four valid checkpoint candidates with highest XID 20 but
aborted with \texttt{APFS object validation failed: checksum mismatch at block
1031} before publishing \texttt{selected\_checkpoint}. The rolling roadmap
\path{docs/research/plans/general-wiztree-for-any-mac-roadmap.md} treats this
as the gating defect for Rust FS-record body decoding and downstream namespace
emission. The next exact step is a focused checkpoint/OMAP-context replay for
the EX-14 fixture shape that distinguishes stale checkpoint selection, a
missing checkpoint-map/data-ring rule, a Rust validation bug, or a
malformed-source hard stop. Until that experiment closes, the manual records
the native pipeline as halted at FS-record family inventory.
