\chapter{The Object Map}

APFS resolves logical object identifiers to physical block addresses
through a B-tree called an object map, or OMAP. There is one OMAP per
container (resolving volume superblocks and ephemerals) and one per
volume (resolving the FS-tree root and its descendants). The OMAP is
the single mechanism by which a copy-on-write filesystem can have
stable object identity over time.

The chapter is short, but its claim is foundational: the OMAP is a
\emph{two-key} lookup, and the second key is non-negotiable.

\section{Lookup By (OID, Max-XID), Not By OID Alone}

\textbf{Spec.} An OMAP key is \texttt{(oid, xid)}. An OMAP value names
a physical address, a size, and a flag word. The OMAP can hold
multiple entries for a single OID, each at a different XID,
corresponding to different versions of the object over time.

\textbf{Spec.} An OMAP lookup specifies an \texttt{oid} and a
\emph{max XID}. The B-tree returns the entry with the highest stored
XID that is \emph{less than or equal to} the requested max XID. The
semantics are lower-bound rather than exact match: the lookup is
asking "give me the version of \texttt{oid} that was current at or
before this scan."

The implication for the caller is structural. The resolver input is
not a bare OID. It is a triple:

\begin{enumerate}
  \item the \emph{OMAP context} --- which OMAP owns the OID,
  \item the OID itself,
  \item the scan XID.
\end{enumerate}

A caller that hands the resolver only an OID is asking the wrong
question. The OID by itself does not determine which version of the
object is meant; the scan XID does.

The scan XID is the same XID for the entire run: the one selected from
the descriptor area in chapter~3. Mixing scan XIDs across object
resolutions within one run reintroduces the incoherence that pinning
the checkpoint was supposed to prevent.

\section{Hard Stops At OMAP Open}

When the parser opens an OMAP --- reads the
\texttt{omap\_phys\_t} root object and validates it --- the
\texttt{omap\_phys.flags} field gates whether the OMAP can be used at
all.

\begin{description}
  \item[\texttt{OMAP\_PHYS\_ENCRYPTING}.]
  \item[\texttt{OMAP\_PHYS\_DECRYPTING}.]
  \item[\texttt{OMAP\_PHYS\_KEYROLLING}.]
  \item[\texttt{OMAP\_PHYS\_CRYPTO\_GENERATION\_FLAG}.]
\end{description}

Any of these set means an encryption operation is in flight on the
container or the volume. Raw mode cannot be supported in such a
state: the bytes the reader sees do not yet match the bytes the
running system will commit. The parser fails closed; the caller
falls back to a supported API.

\textbf{Observation.} Any \emph{unknown} bit in \texttt{omap\_phys.flags}
is also a hard stop. APFS reserves bits over time; an unknown bit
means the running OS knows something the parser does not, and the
parser must not infer behaviour. This is the format-drift discipline
of chapter~11 applied at the smallest object that hosts it.

\section{Hard Stops At OMAP Lookup}

Each value entry returned by an OMAP lookup carries its own flag
word. The flags decide whether the value is usable.

\begin{description}
  \item[\texttt{OMAP\_VAL\_DELETED}.] Negative result, not an error.
    The OID had a mapping at some earlier XID but is no longer
    visible at the scan XID. The caller interprets this as
    "the object is not present" and proceeds accordingly. The
    parser does not hard-stop.
  \item[\texttt{OMAP\_VAL\_ENCRYPTED}.] The value is per-object
    encrypted. Raw mode cannot decrypt; hard stop.
  \item[\texttt{OMAP\_VAL\_NOHEADER}.] The target block does not
    carry an \texttt{obj\_phys\_t} header at all. A structural
    assumption the rest of the parser depends on has been broken;
    hard stop.
  \item[\texttt{OMAP\_VAL\_CRYPTO\_GENERATION}.] Encryption-generation
    mismatch between the OMAP entry and the volume; hard stop.
  \item[unknown bits.] Format drift; hard stop.
\end{description}

The discipline is tighter at lookup time than at open time only
because lookup is also where the parser sees per-entry encryption
behaviour. The principle is the same: known flag means known
behaviour; unknown flag means the parser does not understand the
running system and refuses to guess.

\section{An Empirical Correction: The B-Tree TOC Layout}

The OMAP is a B-tree, and B-tree leaf nodes in APFS carry a key/value
table-of-contents that the reader must traverse to find values. The
TOC format is uniformly described across the Apple reference and the
converged code, but a single offset interpretation gates whether the
reader actually decodes the values correctly. The project landed this
correction during EX-12.

\textbf{Observation.} The TOC \texttt{v\_off} field measures the
distance from the \emph{end} of the value area to the \emph{start}
of the value. The value then runs \emph{forward} from that start
position for \texttt{fixed\_value\_size} bytes (or \texttt{v\_len}
for variable-length values).

The naive reading is "the value extends backward from \texttt{v\_off}
toward the start of the node." Under that reading, OMAP values
decode with a bizarre flag word --- specifically
\texttt{flags=0x10ffff} on the proof fixture --- and the rest of the
parser breaks downstream.

The corrected reading produces clean OMAP values
(\texttt{flags=0x0}, \texttt{size=4096}, \texttt{paddr=...}) and
matches every converged reader. The fix is one constant change in
B-tree value extraction; the regression is one synthetic OMAP test.
The cost of the bug is high (the parser is silently wrong about every
lookup) and the test that catches it is cheap, so the project keeps
the synthetic OMAP test in the regression suite.

\section{Cross-Tool Validation Of The Lookup Contract}

EX-12 paired the project's parser with \texttt{go-apfs identitydump}
against the same \texttt{/dev/rdiskN}. The probe:

\begin{enumerate}
  \item built a fresh proof fixture,
  \item ran the project's parser, recording every resolved
    \texttt{(oid, paddr, xid)} sample published by the OMAP,
  \item replayed object-header validation rules in Python at each
    published \texttt{paddr},
  \item ran \texttt{go-apfs identitydump} against the same raw
    container,
  \item compared the two tools' agreement on the FS-tree root
    \texttt{oid}.
\end{enumerate}

The two agreed on the FS-tree root OID. They did not agree on which
scan XID to select: the project's parser picked the higher XID
visible in the descriptor ring, \texttt{go-apfs} appeared to pick one
lower. That divergence is recorded as a caveat
(\texttt{go\_apfs\_active\_state\_observation}) and not silenced; it
does not affect the value of the cross-tool oracle, because both
tools' OMAP lookups internally use a consistent scan XID and produce
internally consistent root resolutions.

The lesson generalises. Two tools that disagree on selection but agree
on the OID they reach through that selection are still useful as
oracles for each other, as long as the parser records the selection
difference. A cross-tool oracle that silences a known disagreement is
worse than no oracle at all.

\section{What The OMAP Does Not Do}

The OMAP does not resolve objects \emph{within} the FS tree. That is
the source of the next chapter's finding, and the trap that cost the
project a fixture.

FS-tree internal-node entries also carry OIDs, and those OIDs also
require OMAP resolution at the scan XID --- specifically, through the
\emph{volume} OMAP, not the container OMAP. The OMAP machinery from
this chapter is exactly what is needed; what changes is the caller.
The next chapter develops the rule and its empirical evidence.
