\chapter{Validation And Oracles}

\section{Plural Oracles}

There is no single oracle for APFS indexing. The correct oracle depends on the
metric and product mode:

\begin{itemize}
  \item namespace: POSIX/API traversal of the same selected volume or stable view
  \item logical size: public file metadata APIs and raw inode/dstream comparison
  \item allocated size: metric-specific public tool and raw extent evidence
  \item incremental correctness: fresh full raw parse of the same selected state
  \item boot-root semantics: user-visible macOS namespace only when that mode is
  explicitly requested
\end{itemize}

The canonical oracle policy is in
\path{docs/research/RL-10-validation-corpus-and-oracle.md}.

\section{Successful Proof Pattern}

The successful v1 proof pattern is:

\begin{enumerate}
  \item mount a lab image
  \item build a deterministic corpus
  \item capture mounted oracle
  \item detach the image to stabilize raw state
  \item reattach without mounting
  \item raw-walk the APFS container
  \item normalize and compare path, type, file identity, logical size, and
  symlink target
\end{enumerate}

This pattern was established in \path{docs/research/experiments/EX-03-pinned-state-raw-vs-oracle/README.md}
and expanded in \path{docs/research/experiments/EX-04-expanded-pinned-corpus/README.md}.

\section{Cross-Tool Oracle Pairing}

\path{docs/research/experiments/EX-12-omap-lookup-contract/README.md}
established a paired oracle pattern for the Rust path: a probe builds a fresh
proof fixture, runs the native Rust scanner and \texttt{go-apfs identitydump}
against the same \path{/dev/rdiskN}, replays \texttt{obj\_phys\_t} validation at
every Rust-returned \texttt{paddr}, re-runs the SR-006 lower-bound rule on
Rust's published OMAP samples, and confirms cross-tool agreement on
\texttt{root\_tree.oid}. The \texttt{(paddr, object\_xid)} divergence between
the Rust selected XID and the apparent \texttt{go-apfs} active state is recorded
as a \texttt{go\_apfs\_active\_state\_observation} caveat rather than a
contradiction.

\section{Validation Gate Cascade}

The current source-backed validation gates that precede any namespace claim:

\begin{itemize}
  \item \path{docs/research/sources/SR-005-checkpoint-validation-details.md}
  for the checkpoint-superblock validation contract.
  \item \path{docs/research/sources/SR-006-omap-lookup-semantics.md} for
  \texttt{(omap, oid, max\_xid)} lower-bound semantics and OMAP failure cases.
  \item \path{docs/research/sources/SR-007-object-header-fail-closed-validation.md}
  for the single \texttt{validate\_object\_block} chokepoint used by every
  reader in \path{crates/apfs-fastindex}.
  \item \path{docs/research/sources/SR-013-checkpoint-map-integrity.md} for
  checkpoint-map ring validation and ephemeral-object handling.
  \item \path{docs/research/sources/SR-016-record-body-fail-closed-boundary.md}
  for malformed-body hard stops before any row is emitted.
\end{itemize}

\section{Negative Live-State Pattern}

\path{docs/research/experiments/EX-05-live-pinned-churn/README.md} is the
counterexample pattern. A mounted image can be raw-readable while latest-state
raw output still lacks a named stable oracle. That result prevents the manual
from treating live latest raw traversal as correctness evidence.

\section{Inconclusive Pattern}

\path{docs/research/experiments/EX-14-xfield-layout-variant/README.md} is the
counterexample pattern for an upstream failure: the experiment captured a
mounted oracle, ran the Rust scanner, and recorded that the native context
provider failed before xfield comparison could occur. Recording the
\texttt{validation\_gaps} list and the failing block address is itself the
deliverable; the experiment does not promote a Hypothesis past the available
evidence.
