\chapter{The Fail-Closed Boundary}

A parser that emits rows for cases it does not understand is worse than
a parser that emits nothing. The user can react to a missing row by
looking elsewhere; a silently wrong row propagates into aggregates, into
treemaps, and into actions the user takes on the basis of those
aggregates. The cost of being silently wrong is unbounded. The cost of
hard-stopping with a typed reason is bounded: the caller chooses a
fallback, and no false output reaches the user.

Fail-closed is therefore not an aesthetic preference. It is the only
discipline that bounds the cost of incomplete understanding.

\section{Where The Boundary Sits}

The boundary is concentrated in two single chokepoints:

\begin{description}
  \item[Object headers.] One function
    (\texttt{validate\_object\_block} in the Rust crate) validates
    every block before any reader trusts its body. It checks the
    Fletcher-64 checksum, the object type, the expected storage
    class, the encryption flag, the no-header flag, optional
    \texttt{o\_oid == paddr} for physical objects, and the optional
    max-XID guard.
  \item[Record bodies.] One function (\texttt{decode\_fs\_record})
    validates every FS-tree leaf record against the rules below
    before any emission path trusts the decoded values.
\end{description}

There is no second path through either chokepoint. A block that fails
header validation never reaches a body decoder; a record that fails
body validation never reaches the namespace builder. The discipline is
maintained by the topology of the calls, not by every caller
remembering to validate.

A representative hard stop, from the body decoder:

\begin{minted}{rust}
if blob_header.xf_used_data as usize != padded_total {
    return Err(ScanError::InvalidObject(format!(
        "xf_used_data {} disagrees with padded value-area sum {}",
        blob_header.xf_used_data, padded_total,
    )));
}
\end{minted}

The error carries the rule that failed, the observed value, and the
expected value. The caller dispatches on the variant
(\texttt{InvalidObject}), and the substring is preserved for
diagnostic display.

\section{The Per-Record Catalogue}

\textbf{Observation.} Twenty-one per-record fail-closed cases are
enforced by Rust unit tests on synthetic raw blocks. Each test
asserts that the body decoder returns a typed
\texttt{InvalidObject(reason)} with the rule's substring.

\subsection{Short bodies}

\begin{enumerate}
  \item Inode value shorter than the fixed \texttt{j\_inode\_val\_t}
    struct.
  \item Directory-record value shorter than the fixed
    \texttt{j\_drec\_val\_t} struct.
  \item Sibling-link value with \texttt{name\_len} exceeding the bytes
    actually available.
  \item Sibling-map value shorter than 8 bytes.
\end{enumerate}

\subsection{Malformed names}

\begin{enumerate}
  \setcounter{enumi}{4}
  \item Variable-length name length longer than the bytes available.
  \item Embedded NUL inside a required name (other than the
    null-terminator).
  \item Non-UTF-8 bytes in a required name.
  \item Directory-entry type outside the POSIX \texttt{DT\_*}
    allowlist.
\end{enumerate}

\subsection{Malformed xfields}

\begin{enumerate}
  \setcounter{enumi}{8}
  \item Duplicate xfield types inside one blob.
  \item \texttt{xf\_used\_data} disagreeing with
    \texttt{sum(round\_up(x\_size, 8))}.
  \item Xfield value cursor running past the blob.
  \item Xfield metadata table running past the blob.
  \item Xfield blob shorter than the 4-byte \texttt{xf\_blob\_t}
    header.
  \item Required xfield with the wrong \texttt{x\_size}
    (\texttt{DSTREAM} of length other than 40).
  \item Required xfield with the wrong \texttt{x\_size}
    (\texttt{SIBLING\_ID} of length other than 8).
\end{enumerate}

\subsection{Malformed xattrs}

\begin{enumerate}
  \setcounter{enumi}{15}
  \item Xattr value shorter than the fixed header.
  \item Xattr sets both \texttt{EMBEDDED} and \texttt{STREAM} flags.
  \item Xattr sets neither flag.
  \item Xattr sets unknown flag bits.
  \item Xattr body length disagrees with \texttt{xdata\_len}.
  \item Xattr stream body shorter than \texttt{j\_xattr\_dstream\_t}.
\end{enumerate}

\section{The Cross-Record Cases}

Three SR-016 cases depend on consistency across more than one record
and therefore cannot be enforced by synthetic single-record tests.
They are deferred to fixture-level probes; the open questions chapter
tracks their status.

\begin{description}
  \item[Drec type vs inode mode.] A directory entry's
    \texttt{DT\_*} type and the referenced inode's \texttt{S\_IFMT}
    mode must agree. A regular-file inode pointed at by a
    \texttt{DT\_DIR} entry is malformed at the volume level even
    though both records validate in isolation.
  \item[Sibling-map closure.] A directory entry that carries
    \texttt{DREC\_EXT\_TYPE\_SIBLING\_ID} requires a SIBLING\_MAP
    record that maps that sibling ID to the entry's child file ID.
    A missing closure produces a hard-link group with a dangling
    edge.
  \item[Compressed-size precedence conflict.] An inode that sets
    \texttt{INODE\_HAS\_UNCOMPRESSED\_SIZE} but does not carry a
    \texttt{com.apple.decmpfs} xattr is internally inconsistent at
    the cross-record level. Chapter~8 develops the precedence rule
    that prevents the conflict from producing wrong rows.
\end{description}

These cases will close as the project lands EX-22 and the future
cross-record consistency probe.

\section{Why The Boundary Is Tighter Than It Looks}

The fail-closed catalogue may seem heavy for a parser that, in
ordinary operation, encounters none of these cases. The point of the
catalogue is what happens in the not-ordinary case.

Three concrete situations exercise the boundary in practice:

\begin{description}
  \item[Format drift.] APFS reserves bits and types over time. A
    container written by a newer macOS may carry an xattr flag, an
    xfield type, or an incompatible-feature bit the parser does not
    recognise. The unknown-bit hard stops mean a newer container
    produces a typed refusal rather than a quietly wrong row.
  \item[Genuinely malformed images.] Recovery scenarios, partial
    image dumps, or test corpora may include containers with
    structurally inconsistent records. The catalogue refuses them
    with a precise reason, which the caller can present as a
    diagnostic.
  \item[Parser bugs.] If a future change to the body decoder
    silently re-introduces an off-by-one in the xfield cursor, the
    \texttt{xf\_used\_data} structural assertion catches it on the
    first record with a non-multiple-of-eight \texttt{x\_size}. The
    catalogue is also a regression suite.
\end{description}

The catalogue is therefore proportional to the cost it prevents, not
to the frequency with which it is exercised.

\section{The Object-Header Side}

The header-validation chokepoint is smaller but no less load-bearing.
A block that does not validate is never decoded; a block that
validates under the wrong storage class is also rejected, because
\texttt{validate\_object\_block} takes the expected storage as a
parameter.

The storage classes the parser uses:

\begin{description}
  \item[\texttt{Physical}.] The object's \texttt{o\_oid} equals its
    physical block address. Container superblocks, checkpoint maps,
    OMAP nodes, and free-queue nodes are physical.
  \item[\texttt{Virtual}.] The object lives in a virtual address
    space resolved by an OMAP. \texttt{o\_oid} is independent of the
    block's paddr; the validator checks
    \texttt{o\_oid == omap\_lookup.oid} instead. Volume superblocks
    (looked up through the container OMAP), FS-tree roots and
    internal nodes (looked up through the volume OMAP), and dstream
    extent nodes are virtual.
  \item[\texttt{Ephemeral}.] In-memory between transactions; reached
    through the checkpoint map. The parser uses the checkpoint map's
    mapping, never an ephemeral storage flag.
\end{description}

Mixing classes is a structural mistake. Chapter~3 records the empirical
correction that checkpoint-map \emph{blocks} are themselves physical,
not ephemeral, despite naming objects that are ephemeral; chapter~5
records that FS-tree roots are virtual despite living at a known
paddr that the volume OMAP returns. The parser keeps both rules
isolated to their call sites.

\section{The Boundary And The Fallback}

The fail-closed boundary is the upstream half of the support contract;
the fallback path (chapter~11) is the downstream half. When the
boundary refuses a source, the caller is not stuck: it dispatches to
the POSIX-traversal fallback, which produces the same shape of output
through supported APIs. The user sees an answer; the wrong answer is
not produced.

This is the design pattern the rest of the manual presupposes: known
behaviour is enforced as a hard rule, unknown behaviour is escalated
to a path that asks the operating system rather than guessing on its
behalf.
