\chapter{Sizes and Their Precedences}

APFS stores file size in several places, and the places do not
always agree. The parser emits three per-inode metrics, each with
its own precedence rule and its own oracle:

\begin{description}
  \item[Logical.] The user-visible byte length (\texttt{st\_size}).
    The primary anchor; covers every case.
  \item[Allocated.] Bytes APFS has assigned to the inode's content
    (\texttt{st\_blocks * 512}). Covers every shape the macOS
    write path produces.
  \item[Real.] Clone-deduplicated allocated bytes. Each inode is
    charged its share of physically-allocated bytes, accounting for
    extents that multiple inodes reference. Validated by the
    extent-reference tree.
\end{description}

The chapter develops them in that order.

\section{Logical Size}

Logical size is the user-visible byte length: the number POSIX
\texttt{st\_size} reports and the number \texttt{ls -l} prints. It
has a stable oracle, applies uniformly to ordinary, sparse, cloned,
hard-linked, symlink, and compressed files, and is the size every
user-facing tool agrees on.

\subsection{The Precedence Rule}

\textbf{Observation.} Five steps, applied per inode, produce
\texttt{st\_size} for every case the validated corpus covers:

\begin{enumerate}
  \item If the inode's \texttt{internal\_flags} carry
    \texttt{INODE\_HAS\_UNCOMPRESSED\_SIZE}: pick
    \texttt{inode.uncompressed\_size}.
  \item Else if the inode carries a \texttt{j\_dstream\_t} (via
    \texttt{INO\_EXT\_TYPE\_DSTREAM} or a paired
    \texttt{j\_dstream\_id\_val\_t} record): pick
    \texttt{j\_dstream\_t.size}.
  \item Else if a \texttt{com.apple.decmpfs} xattr is present and
    step 1 did not fire: pick the decmpfs header's
    \texttt{uncompressed\_size} field.
  \item For symlinks (\texttt{S\_IFLNK}): pick the byte length of the
    \texttt{com.apple.fs.symlink} xattr payload.
  \item Otherwise: zero.
\end{enumerate}

The order matters. Steps 1 and 3 both reach into compression
metadata, but they reach for different fields and have different
trust; the load-bearing case below shows why.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  every node/.style={font=\footnotesize},
  decision/.style={diamond, draw, aspect=2.4, inner sep=1pt,
                   align=center, fill=blue!8},
  result/.style={draw, rounded corners=2pt, fill=green!18,
                 minimum width=3.6cm, minimum height=0.8cm, align=center},
  fallback/.style={draw, rounded corners=2pt, fill=gray!15,
                   minimum width=3.0cm, minimum height=0.8cm, align=center},
  flow/.style={-Stealth},
  yeslab/.style={font=\scriptsize, midway, above, sloped=false,
                 fill=white, inner sep=1pt},
  nolab/.style={font=\scriptsize, midway, right=2pt,
                fill=white, inner sep=1pt},
]
  \node[decision] (d1) {INODE\_HAS\_\\UNCOMPRESSED\_SIZE?};
  \node[result, right=2.0cm of d1] (r1) {\texttt{inode.uncompressed\_size}};

  \node[decision, below=1.2cm of d1] (d2) {\texttt{j\_dstream\_t}\\present?};
  \node[result, right=2.0cm of d2] (r2) {\texttt{j\_dstream\_t.size}};

  \node[decision, below=1.2cm of d2] (d3) {decmpfs xattr\\present?};
  \node[result, right=2.0cm of d3] (r3) {decmpfs header\\\texttt{uncompressed\_size}};

  \node[decision, below=1.2cm of d3] (d4) {symlink\\(S\_IFLNK)?};
  \node[result, right=2.0cm of d4] (r4) {symlink target\\byte length};

  \node[fallback, below=1.2cm of d4] (r5) {zero};

  \draw[flow] (d1.east) -- (r1.west) node[yeslab] {yes};
  \draw[flow] (d2.east) -- (r2.west) node[yeslab] {yes};
  \draw[flow] (d3.east) -- (r3.west) node[yeslab] {yes};
  \draw[flow] (d4.east) -- (r4.west) node[yeslab] {yes};

  \draw[flow] (d1.south) -- (d2.north) node[nolab] {no};
  \draw[flow] (d2.south) -- (d3.north) node[nolab] {no};
  \draw[flow] (d3.south) -- (d4.north) node[nolab] {no};
  \draw[flow] (d4.south) -- (r5.north) node[nolab] {no};
\end{tikzpicture}
\caption{Per-inode logical-size precedence. The order is non-negotiable;
  the compressed case is the load-bearing reason step~1 must beat
  step~3.}
\end{figure}

In pseudocode the same rule reads as one short function:

\begin{minted}{python}
def logical_size(inode, xattrs):
    if inode.internal_flags & INODE_HAS_UNCOMPRESSED_SIZE:
        return inode.uncompressed_size
    if inode.dstream is not None:
        return inode.dstream.size
    decmpfs = xattrs.get("com.apple.decmpfs")
    if decmpfs is not None:
        return decmpfs.header.uncompressed_size
    if S_ISLNK(inode.mode):
        return len(xattrs["com.apple.fs.symlink"].payload)
    return 0
\end{minted}

\subsection{The Load-Bearing Case: Compression}

\textbf{Observation.} On the EX-19 fixture the compressed file
(\texttt{compressed.txt}, created via
\texttt{ditto --hfsCompression}) presents:

\begin{itemize}
  \item \texttt{INODE\_HAS\_UNCOMPRESSED\_SIZE} set,
    \texttt{inode.uncompressed\_size = 53248}.
  \item No \texttt{j\_dstream\_t}. APFS stores small compressed data
    inline in the decmpfs xattr and does not allocate a separate
    dstream for it.
  \item \texttt{com.apple.decmpfs} xattr present, with header
    \texttt{uncompressed\_size = 437} and
    \texttt{compression\_type = 0}.
  \item Mounted POSIX \texttt{st\_size = 53248}.
\end{itemize}

The decmpfs header's \texttt{uncompressed\_size} is a placeholder
for this layout. Step~1 must beat step~3 or the parser emits 437
instead of 53248. The principle is general: when APFS sets the
inode flag, the inode flag is the source of truth; the decmpfs
header is metadata the running system maintains for its own
decompression path and is not reliably the file's size.

\subsection{Sparse Bytes Is A Hint, Not A Size}

\textbf{Observation.}
\texttt{INO\_EXT\_TYPE\_SPARSE\_BYTES} is an
\emph{allocation hint}, not a logical size. It records the bytes the
sparse-allocation layer has not committed to disk.

On the proof fixture's sparse file:

\begin{itemize}
  \item \texttt{j\_dstream\_t.size = 1052897} (which equals
    \texttt{st\_size}).
  \item \texttt{INO\_EXT\_TYPE\_SPARSE\_BYTES = 1032192}.
  \item \texttt{st\_size = 1052897}.
\end{itemize}

\texttt{SPARSE\_BYTES} records the difference between the logical
extent and the allocated bytes. It is the load-bearing input to the
allocated-size rule (next section); it is not the answer to ``how
large is this file.'' Treating \texttt{SPARSE\_BYTES} as anything
else inflates the logical size of every sparse file.

\subsection{Per-Inode Results}

\textbf{Observation.} On the EX-19 same-run fixture covering ordinary,
sparse, clone, hard-linked, symlink, and compressed cases, the
precedence rule reproduces \texttt{st\_size} for every unique inode:

\begin{center}
\begin{tabular}{lrlrr}
\toprule
path & inode & step picked & value & st\_size \\
\midrule
ordinary.txt   & 16 & \texttt{j\_dstream\_size}         & 29      & 29 \\
sparse.bin     & 19 & \texttt{j\_dstream\_size}         & 1052897 & 1052897 \\
clone.txt      & 20 & \texttt{j\_dstream\_size}         & 29      & 29 \\
link.txt       & 23 & \texttt{symlink\_target\_len}     & 12      & 12 \\
compressed.txt & 24 & \texttt{inode\_uncompressed\_size} & 53248   & 53248 \\
\bottomrule
\end{tabular}
\end{center}

Zero mismatches. The compressed case picks step~1; everything else
picks step~2 except the symlink (step~4). The hard link does not
appear as a separate row because the rule is per inode and
\texttt{hard.txt} shares inode~16 with \texttt{ordinary.txt}.

\section{Allocated Size}

Per inode, the bytes APFS has assigned to the inode's content
stream. Oracle: \texttt{st\_blocks * 512}.

\textbf{Observation.} The rule covers every shape the macOS write
path produces. Ordinary, cloned, hard-linked, and symlink cases
emit \texttt{j\_dstream\_t.alloced\_size} (zero for symlinks). The
sparse and decmpfs cases were promoted from fail-closed to
oracle-validated by EX-26: sparse emits
\texttt{alloced\_size - sparse\_bytes}; decmpfs emits the sum of
any stream-backed xattrs' \texttt{alloced\_size} (the primary
dstream is usually absent; embedded xattrs contribute zero because
their bytes live inline in the xattr payload).

\subsection{The Precedence Rule}

\textbf{Observation.} Six branches, applied per inode:

\begin{enumerate}
  \item Regular file with a \texttt{j\_dstream\_t} xfield and no
    \texttt{INO\_EXT\_TYPE\_SPARSE\_BYTES} xfield and no
    \texttt{com.apple.decmpfs} xattr: emit
    \texttt{Some(j\_dstream\_t.alloced\_size)}.
  \item Regular file with a \texttt{j\_dstream\_t} xfield and a
    \texttt{SPARSE\_BYTES} xfield (no decmpfs): emit
    \texttt{Some(alloced\_size - sparse\_bytes)} (EX-26 Hypothesis~A,
    see §\ref{ssec:sparse-bytes-lift}).
  \item Regular file with a \texttt{com.apple.decmpfs} xattr: emit
    \texttt{Some(primary + decmpfs.stream\_dstream.alloced +
    ResourceFork.stream\_dstream.alloced)} (each term defaults to
    zero when absent or embedded; EX-26 Hypothesis~F, see
    §\ref{ssec:decmpfs-lift}).
  \item Symlinks: emit \texttt{Some(0)}.
  \item Directories: emit \texttt{Some(0)}.
  \item Anything else: emit \texttt{None}.
\end{enumerate}

\texttt{None} is not a synonym for zero. A row with
\texttt{allocated\_size = None} is recorded in the run's
\texttt{not\_claimed} register; the metric toggle in the
visualisation greys out for that row; any directory aggregate
containing a \texttt{None}-allocated inode is itself emitted as
\texttt{None}. A partial aggregate is forbidden, since a consumer
might otherwise treat it as authoritative.

\subsection{Why \texttt{alloced\_size} Is The Only Candidate}

A survey of six independent open-source readers
(\texttt{linux-apfs-rw}, \texttt{apfsprogs}/\texttt{apfsck},
the Sleuth Kit, \texttt{libfsapfs}, \texttt{dissect.apfs},
\texttt{apfs-fuse}, \texttt{go-apfs}) finds all six decode
\texttt{j\_dstream\_t.alloced\_size}; only TSK and
\texttt{linux-apfs-rw} surface it to callers; no reader surfaces
the alternative candidate (the sum of
\texttt{j\_file\_extent\_val\_t.len}) as a public allocated-bytes
number. The extent-reference tree is the per-volume
reference-counting authority for clones and snapshots, not a
per-file allocated-bytes table; it is used here for the Real metric
below.

The two halves of the same upstream project disagree:
\texttt{linux-apfs-rw} (kernel) writes
\texttt{alloced\_size = round\_up(ds\_size, blocksize)};
\texttt{apfsck} (filesystem checker) enforces
\texttt{alloced\_size == sum(j\_file\_extent\_val\_t.len)}. The two
agree only when every extent ends at \texttt{ds\_size} and the file
has no truncate residue. The right oracle is the external public
number (\texttt{st\_blocks * 512}), not internal field consistency.

\subsection{The Sparse Identity (EX-22, EX-26)}
\label{ssec:sparse-bytes-lift}

EX-22 left sparse fail-closed because the fixture carried one
sparse shape and the observed identity
\texttt{alloced\_size - sparse\_bytes = st\_blocks * 512} needed
corpus-level confirmation. EX-26 ran the probe on four sparse
shapes; the identity held exactly across all of them.

\textbf{Observation.} On the EX-22 fixture the sparse file carries:

\begin{itemize}
  \item \texttt{j\_dstream\_t.size = 1052897} (equals \texttt{st\_size}).
  \item \texttt{j\_dstream\_t.alloced\_size = 1056768}
    ($= \text{round\_up}(\text{size}, 4096)$).
  \item \texttt{INO\_EXT\_TYPE\_SPARSE\_BYTES = 1032192}
    (the unallocated bytes inside the logical extent).
  \item \texttt{st\_blocks = 48}, so
    \texttt{st\_blocks * 512 = 24576}.
\end{itemize}

\texttt{alloced\_size} overstates the actual on-disk allocation by
exactly $1{,}056{,}768 - 24{,}576 = 1{,}032{,}192$ bytes --- the
sparse-bytes hint. APFS on macOS follows the
\texttt{linux-apfs-rw} kernel rule rather than the
\texttt{apfsck} sum-of-extents rule. The emission rule must
therefore split the dstream case by presence of the
\texttt{SPARSE\_BYTES} xfield.

\textbf{EX-26 corpus result.} The identity
\texttt{alloced\_size - sparse\_bytes = st\_blocks * 512} holds
exactly across four sparse shapes:

\begin{center}
\begin{tabular}{lrrrr}
\toprule
shape & \texttt{alloced\_size} & \texttt{sparse\_bytes} & difference & \texttt{st\_blocks * 512} \\
\midrule
1 MiB HEAD/TAIL hole          & 1{,}056{,}768  & 1{,}032{,}192  &    24{,}576 &    24{,}576 \\
10 MiB hole                   & 10{,}485{,}760 & 10{,}452{,}992 &    32{,}768 &    32{,}768 \\
50 MiB hole                   & 52{,}428{,}800 & 52{,}396{,}032 &    32{,}768 &    32{,}768 \\
4 KiB-data-per-64 KiB chunked & 2{,}097{,}152  & 1{,}572{,}864  &   524{,}288 &   524{,}288 \\
\bottomrule
\end{tabular}
\end{center}

The branch now emits
\texttt{Some(alloced\_size.saturating\_sub(sparse\_bytes))}, with
saturation guarding the (not observed in practice but algebraically
possible) case where the kernel's residue exceeds the dstream-
reported allocation.

\subsection{The decmpfs Rule (EX-26)}
\label{ssec:decmpfs-lift}

EX-22 left decmpfs fail-closed because no open-source reader
agreed on the macOS oracle and the fixture carried only one
decmpfs shape. EX-26 added a 256~KiB compressible payload that
ditto routes into the resource-fork-stored shape. Both shapes
satisfy one rule: allocated bytes equal the sum of any
stream-backed xattrs' \texttt{alloced\_size} plus the (typically
absent) primary dstream \texttt{alloced\_size}.

\textbf{Observation.} On the EX-26 fixture the two decmpfs files
carry:

\begin{itemize}
  \item \texttt{compressed.txt} (52~KiB compressible JSON): no
    primary dstream; \texttt{com.apple.decmpfs} is
    \emph{stream-backed} (flags = \texttt{0x01}) with
    \texttt{stream\_dstream.alloced\_size = 4096};
    \texttt{com.apple.ResourceFork} is stream-backed with
    \texttt{stream\_dstream.alloced\_size = 0} (empty fork).
    Sum: 4096. Oracle \texttt{st\_blocks * 512 = 4096}.
  \item \texttt{compressed-big.bin} (256~KiB compressible binary):
    no primary dstream; \texttt{com.apple.decmpfs} is
    \emph{embedded} (flags = \texttt{0x02}) with payload
    \texttt{fpmc} magic, \texttt{compression\_type = 8} (lzvn
    resource-fork-stored), \texttt{uncompressed\_size = 262144};
    \texttt{com.apple.ResourceFork} is stream-backed with
    \texttt{stream\_dstream.alloced\_size = 4096}. Sum: 4096.
    Oracle \texttt{st\_blocks * 512 = 4096}.
\end{itemize}

The decmpfs header's \texttt{compression\_type} byte tells the
reader \emph{where} the compressed bytes live (xattr inline vs.\
resource fork). The bytes themselves are always counted by summing
the \texttt{stream\_dstream.alloced\_size} fields of every xattr
that is stream-backed. Embedded xattrs contribute zero because
their bytes live inline in the xattr payload, which is part of the
inode body's storage and does not allocate separate extents.

The disagreement EX-22 documented between \texttt{linux-apfs-rw}
(publishes uncompressed size / 512) and \texttt{apfs-fuse}
(publishes zero) resolves against the macOS oracle: neither is
right; the sum-of-stream-dstreams formula reproduces
\texttt{st\_blocks * 512} exactly.

The incompressible-payload shape, where ditto leaves
\texttt{UF\_COMPRESSED} off and emits no \texttt{com.apple.decmpfs}
xattr, falls through to branch~1 and emits
\texttt{Some(alloced\_size)} under the EX-22 baseline. The EX-26
\texttt{compressed-random.bin} case does exactly this.

\subsection{Per-Inode Results}

\textbf{Observation.} On the EX-26 fixture, all ten rows match the
\texttt{st\_blocks * 512} oracle exactly:

\begin{center}
\begin{tabular}{lllrr}
\toprule
path & branch & rule & picked & st\_blocks * 512 \\
\midrule
ordinary.txt          & 1 & \texttt{alloced\_size}                                              &      4096 &      4096 \\
clone.txt             & 1 & \texttt{alloced\_size}                                              &      4096 &      4096 \\
compressed-random.bin & 1 & \texttt{alloced\_size}                                              &  131{,}072 &  131{,}072 \\
sparse.bin            & 2 & \texttt{alloced\_size - sparse\_bytes}                              &     24{,}576 &     24{,}576 \\
sparse-medium.bin     & 2 & \texttt{alloced\_size - sparse\_bytes}                              &     32{,}768 &     32{,}768 \\
sparse-large.bin      & 2 & \texttt{alloced\_size - sparse\_bytes}                              &     32{,}768 &     32{,}768 \\
sparse-chunked.bin    & 2 & \texttt{alloced\_size - sparse\_bytes}                              &    524{,}288 &    524{,}288 \\
compressed.txt        & 3 & \texttt{decmpfs.stream\_dstream.alloced} (xattr-stream-stored)      &      4096 &      4096 \\
compressed-big.bin    & 3 & \texttt{ResourceFork.stream\_dstream.alloced} (resource-fork-stored)&      4096 &      4096 \\
link.txt              & 4 & \texttt{zero} (symlink)                                              &         0 &         0 \\
\bottomrule
\end{tabular}
\end{center}

Zero mismatches; every row picks a value the rule emits, against
the macOS oracle. The hard-linked file \texttt{hard.txt} does not
appear as a separate row because it shares its inode with
\texttt{ordinary.txt} and contributes through the same per-inode
rule.

\section{Real (Clone-Deduplicated) Size}
\label{ssec:real-size}

Allocated size charges each clone its full dstream. The volume's
on-disk allocation is smaller: clones share extents, and a survey of
a clone-heavy directory using allocated bytes overstates the disk
usage. The \emph{real} size charges each inode its share of
physically-allocated bytes, so a directory total equals the unique
on-disk bytes in its subtree.

\textbf{Oracle.} Per-inode \texttt{du -A} on macOS reports
apparent size (it reads \texttt{st\_blocks * 512} via
\texttt{lstat}, which does not dedup clones), so there is no
public per-file macOS oracle. The on-disk authority is the
extent-reference tree itself:
$\sum \texttt{phys\_ext.length} \cdot \texttt{block\_size}$ is the
ground truth for the volume's allocated bytes.

\subsection{The Formula (EX-27)}

Three APFS trees feed the computation: the FS-tree's \texttt{inode}
records, the FS-tree's \texttt{file\_extent} records (keyed by
\texttt{dstream\_id} --- clones share a dstream, not an inode), and
the extent-reference tree's \texttt{phys\_ext} records (one per
shared physical extent, carrying \texttt{(paddr\_first,
length\_blocks, refcnt, owning\_obj\_id)}).

\begin{enumerate}
  \item For each \texttt{file\_extent}, split its physical byte
    range against the overlapping \texttt{phys\_ext} records. A
    \texttt{phys\_ext} record's \texttt{refcnt} is the count of
    \texttt{file\_extent}s pointing at that range; sub-extents
    not covered by any \texttt{phys\_ext} record have implicit
    \texttt{refcnt = 1}.
  \item Per dstream:
    $\texttt{dstream\_dedup} = \sum_i
    \texttt{sub\_extent}_i\texttt{.length} /
    \texttt{sub\_extent}_i\texttt{.refcnt}$.
  \item Per inode:
    $\texttt{real\_size} = \texttt{dstream\_dedup} /
    \texttt{dstream.refcnt}$,
    where \texttt{dstream.refcnt} is the count of inodes sharing
    this dstream (from the \texttt{j\_dstream\_id\_val\_t} record).
\end{enumerate}

\textbf{Invariants.} On the EX-27 fixture (three clone families:
5-way, 3-way, and a partial-share rewrite):

\begin{itemize}
  \item $\sum_\text{dstreams}\texttt{dstream\_dedup} =
    \sum_\text{phys\_ext}\texttt{length\_blocks} \cdot
    \texttt{block\_size}$ exactly.
  \item $\sum_\text{inodes}\texttt{real\_size}$ equals the same
    total modulo integer-division rounding of $\leq
    (\texttt{refcnt} - 1)$ bytes per clone-shared dstream
    (2-byte residue on the EX-27 fixture across two shared
    dstreams).
\end{itemize}

The second invariant is what makes the metric sum cleanly across
the treemap: per-clone shares summed over a directory equal the
unique on-disk bytes in that subtree.

\subsection{Storage Class Of The Extent-Reference Tree}

\textbf{Observation.} On detached \texttt{.dmg} images
\texttt{hdiutil} creates, the extent-reference tree carries
\texttt{OBJECT\_TYPE\_PHYSICAL} storage --- the OID is the root
paddr directly. Live volumes can use virtual storage, in which case
the volume OMAP returns the root paddr. The crate reads
\texttt{extentref\_tree\_type\_raw} on the volume superblock and
dispatches; a virtual tree with no OMAP mapping at the scan XID
records a validation gap.

\subsection{Fallback Behaviour}

The POSIX-traversal fallback has no access to per-extent
refcounts. It emits \texttt{real\_size = allocated\_size}: the
fallback walker cannot dedup, and saying so is more honest than
publishing the wrong number under the same column heading.

\section{Directory Aggregates}

\textbf{A directory's total is the \emph{unique-inode logical sum
within the aggregate root}.}

A regular file's logical size contributes to a directory total once,
regardless of how many hard-linked paths within that directory tree
reference the same inode. A symlink contributes its target byte
length to its own row but contributes nothing to any directory total
(the target may live elsewhere or not exist at all, and even when it
exists its size is already counted at the target's row).

\textbf{Why this policy.} The naive policy --- sum the logical size of
every path under a directory --- catastrophically over-counts on real
directories with hard-linked content: mail spools, package caches,
Git object stores, snapshot archives. The user looking at a 12 GB
directory deserves to know that 11 GB of it is a single 11 GB file
that also appears elsewhere, not that the directory legitimately
holds 12 GB of unique content. WizTree-on-NTFS rarely encounters
hard-link density that exposes this; on APFS the case is common.

\textbf{Consequence.} Sibling directory totals are not strictly
additive into their parent when the same file identity appears in
more than one subtree. This is correct, not buggy, but the user
interface must label aggregate semantics so the user understands
what the number means.

The aggregate rule is implemented in one place
(\texttt{aggregate.py}, ported to the Rust crate's
\texttt{namespace} module) using
\texttt{setdefault(file\_id, logical\_size)} keyed by inode --- the
policy is enforced by construction rather than by careful summing.

Allocated-size aggregates use the same unique-inode rule with a
\texttt{None}-collapse step: any contributing inode whose
\texttt{allocated\_size} is \texttt{None} (any case not covered by
the EX-26 rules) propagates \texttt{None} up to the directory
total. Real-size aggregates follow the same shape: per-inode
shares sum across the directory, and a \texttt{None} contribution
(from the fallback walker, or from a raw run where the
extent-reference tree did not validate) collapses the total.

\section{Out Of Scope}

Two metrics remain outside the chapter's defended scope; chapter~13
records each as an open question.

\begin{description}
  \item[Exclusive size.] Bytes owned only by this inode after
    accounting for clones \emph{and} snapshots. The chapter's Real
    metric handles clones (extent-reference tree); the
    snapshot-metadata tree must be joined for full exclusivity.
  \item[Shared size.] Bytes shared with at least one other inode
    or snapshot. The complement of exclusive.
\end{description}
