\chapter{Experiment Catalog}

\begin{longtable}{p{0.16\textwidth}p{0.72\textwidth}}
\textbf{Experiment} & \textbf{Purpose} \\
\hline
\texttt{EX-01} & Live checkpoint consistency and runtime boundary. Shows that
mounted lab images are useful probes but latest checkpoint moves under writes. \\
\texttt{EX-02} & Required-record taxonomy and namespace/size corpus design. \\
\texttt{EX-03} & First pinned-state raw-vs-oracle proof loop. \\
\texttt{EX-04} & Expanded pinned corpus across case-insensitive and
case-sensitive image-backed APFS. \\
\texttt{EX-05} & Live mounted churn probe. Shows current latest-state raw walker
does not validate against baseline or final mounted oracles. \\
\texttt{EX-06} & OID, paddr, XID, checksum, and block-hash identity tracking. \\
\texttt{EX-07} & Subtree reuse proof probe for exact node-identity summary reuse. \\
\texttt{EX-08} & Read-path support matrix. First safe-host execution supports
detached unencrypted image-backed APFS for narrow v1 proof work, keeps mounted
images \texttt{readable\_not\_supported}, and records startup raw read as
\texttt{blocked\_privilege}. \\
\texttt{EX-09} & Accounting probe design for logical, allocated, shared,
exclusive, compression, and snapshot-retained semantics. \\
\texttt{EX-10} & Native Rust checkpoint scanner. Source gating, descriptor
scan, NX superblock decode, checkpoint-map validation, container/volume OMAP
\texttt{(oid, max\_xid)} resolution, volume superblock decode under the v1
allowlist, FS-tree root validation, and FS-record family dump. \\
\texttt{EX-11} & Checkpoint-map integrity. Validates the checkpoint-map chain
and mapped ephemeral objects on a generated proof fixture and on synthetic
malformed checkpoint-map hard-stop cases before native OMAP/root traversal. \\
\texttt{EX-12} & OMAP lookup contract. Self-paired probe builds a fresh proof
fixture, runs the native Rust scanner and \texttt{go-apfs identitydump} against
the same \path{/dev/rdiskN}, replays \texttt{obj\_phys\_t} validation at every
Rust-returned \texttt{paddr}, re-runs SR-006 lower-bound on Rust's published
OMAP samples, and confirms cross-tool agreement on \texttt{root\_tree.oid}.
Verdict: \texttt{validated\_omap\_lookup\_contract} for the proof fixture. \\
\texttt{EX-13} & Native FS-record body oracle (Python-first). Decoded
\texttt{DIR\_REC}, \texttt{INODE}, \texttt{XATTR}, \texttt{SIBLING\_LINK},
\texttt{SIBLING\_MAP}, and dstream candidates; reconstructed all mounted paths;
and selected the blob-relative xfield data alignment matching
\texttt{j\_dstream\_t.size} and \texttt{INO\_EXT\_TYPE\_SPARSE\_BYTES}. Verdict:
\texttt{validated\_native\_record\_body\_contract}; four non-row-critical
records still have top-score xfield-layout ambiguity, motivating the
fixture-variant successor. \\
\texttt{EX-14} & Xfield layout variant. Returned
\texttt{oracle\_inconclusive}: Rust scanner found four valid checkpoint
candidates with highest XID 20 but aborted on
\texttt{APFS object validation failed: checksum mismatch at block 1031} before
publishing \texttt{selected\_checkpoint}, blocking xfield comparison and Rust
FS-record body decoding. Treated as a checkpoint/OMAP-context blocker rather
than an FS-record-body finding. \\
\end{longtable}

Each experiment's durable conclusion is in
\path{docs/research/experiments/EX-*/README.md}. Generated JSON and command
outputs belong under the corresponding \path{artifacts/} directory. Negative
and inconclusive experiments are catalog entries on the same footing as
positive ones.
