\chapter{Names Across Volume Modes}

A row in the namespace is identified by a path, and a path is a
sequence of names. The names come from directory record keys; the
keys are stored as UTF-8 byte strings. The chapter develops one
finding and one deliberate non-promise.

\section{The Finding}

\textbf{Observation.} Stored names pass through verbatim.

The Rust body decoder emits the bytes the directory key stores. The
parser does not normalise, case-fold, or interpret. POSIX traversal
of the same volume returns the same bytes. The two agree byte for
byte.

The finding is stated in the simplest form on purpose. Anything more
elaborate --- "Rust emits the canonical form" or "Rust converts to
NFC" --- would be wrong on some volume and would silently rename
files in the output. Emit what the volume stored; the user can
compare what they see to what \texttt{ls} shows.

\section{Two Volume Modes Exist}

APFS volumes come in two case-and-normalisation modes that v1 cares
about.

\begin{description}
  \item[\texttt{APFS} (case-insensitive).]
    \texttt{case\_insensitive = true},
    \texttt{normalization\_insensitive = false}. The volume's
    incompatible-feature mask sets the case-insensitive bit only.
    The volume is also normalisation-insensitive for lookup, but
    APFS reserves the explicit incompat bit for the case-sensitive
    variant. The asymmetry is a quirk of the on-disk format, not a
    statement about behaviour.
  \item[\texttt{Case-sensitive APFS}.]
    \texttt{case\_insensitive = false},
    \texttt{normalization\_insensitive = true}.
\end{description}

Both modes appear on real systems, and both have to be supported.

\section{What The Modes Do Affect}

The modes affect \emph{lookup}: when the kernel resolves a path, it
consults the volume's case and normalisation behaviour to decide
whether two names refer to the same file. They do not affect row
\emph{enumeration}: enumeration walks directory keys and returns
their bytes.

The distinction is the chapter's central claim. The parser can emit
correct rows on both modes without performing the lookup-equivalence
test, because the lookup behaviour is the running system's
responsibility, not the parser's.

\subsection{The Lookup Oracle}

\textbf{Observation.} Across the two volume modes, APFS lookup
behaves as follows. On the EX-20 same-run fixture using
\texttt{os.O\_CREAT | os.O\_EXCL}:

\begin{center}
\begin{tabular}{lll}
\toprule
operation & CI volume & CS volume \\
\midrule
create \texttt{plain.txt}                          & accept & accept \\
create \texttt{CaseName.txt}                       & accept & accept \\
\texttt{casename.txt} after \texttt{CaseName.txt}  & EEXIST & accept \\
create NFC \texttt{caf\'e.txt}                     & accept & accept \\
NFD \texttt{caf\'e.txt} after NFC sibling          & EEXIST & EEXIST \\
\bottomrule
\end{tabular}
\end{center}

The case-insensitive volume rejects the lowercase form because lookup
considers the two names equivalent. Both volumes reject the
decomposed Unicode form after the precomposed sibling because both
treat NFC and NFD as the same name for lookup. This is what APFS does;
no APFS-specific knowledge in the indexer is needed to produce the
result, because the kernel did the work.

\subsection{Rust Vs POSIX Equality}

\textbf{Observation.} On both volumes the Rust-emitted path bytes
match POSIX-traversal path bytes exactly. The comparison is byte for
byte: \texttt{path.encode("utf-8").hex()} produces the same string on
both sides for every entry. On the CI volume the corpus has four
entries; on the CS volume it has five (the case-distinct sibling
also appears); the equality holds for every one of them.

The cleanest way to see this: both sides walk the directory key
storage. POSIX walks through the kernel; Rust walks through the FS
tree directly. Neither side applies normalisation, and the bytes are
the same bytes either way.

\section{The Deliberate Non-Promise}

The parser does not implement APFS lookup-by-name.

The APFS lookup primitive is the directory hash: a CRC32C over the
name's UTF-32LE codepoints, with NFD normalisation applied to the
input first, and on case-insensitive volumes Unicode case-fold
applied before normalisation. Two names with the same hash collide
within a directory; APFS resolves the collision by storing both keys
and walking them.

A \emph{find this exact path} feature has to reproduce that hash to
locate the directory entry. The project has not yet validated a hash
implementation against a same-run hash oracle, and until it does,
the parser does not pretend to find files by name.

The asymmetry is deliberate. A row enumeration that is correct ---
every visible path appears once, with the right bytes --- is useful
immediately: it powers a treemap, sort-by-size, drill-down, and
context-menu operations on rows the user has selected. A
lookup-by-name that is silently wrong returns the user's query
against the wrong file. The cost of being wrong about a lookup is
unbounded; the cost of refusing to do the lookup is "the user has
to navigate to the file rather than type its name." The asymmetry
of cost dictates the asymmetry of scope.

\section{Names And Aggregates}

The name preservation rule has one further consequence. Directory
aggregates (chapter~8) key on inode identity, not on path. Two paths
that hard-link to the same inode contribute that inode's logical
size once to a directory total, regardless of whether their names
are casewise or normalisationally different.

The hard-link unification reads SIBLING\_LINK and SIBLING\_MAP
records to find every name that points at a given inode. The names
are emitted verbatim in each row; the inode is recorded as the row's
\texttt{file\_id}; the aggregate keys on the file\_id. Names and
sizes are decoupled from aggregate identity, which is the right
shape for a filesystem that allows the same content to live under
many names.

\section{What Future Work Will Need}

A future lookup-by-name feature will need:

\begin{enumerate}
  \item A same-run hash oracle: a probe that takes a constructed
    name, runs APFS's own hash through a controlled API (the
    kernel's \texttt{O\_EXCL} accept/reject is the cheapest
    indicator), and produces a positive case and a negative case
    for the hash equivalence class.
  \item NFD normalisation tables and Unicode case-fold tables that
    match the macOS version the volume was written on, accounting
    for tables changing across Unicode revisions.
  \item Collision handling: directories may store multiple keys with
    the same hash, and the lookup must walk all of them.
  \item Volume-mode dispatch: case-insensitive volumes apply case-fold
    before normalisation; case-sensitive volumes do not.
\end{enumerate}

That feature has its own oracle and its own probe sequence and is
out of v1. The non-promise here is what keeps v1's namespace claim
defensible.
