\chapter{Identity, Reuse, and Cache Keys}

A second scan of an unchanged volume should be faster than the
first. The natural mechanism is a subtree-summary cache: walk the
FS tree once, summarise each node, reuse the summary on the next
run when the node has not changed. The chapter develops what
identity that cache must key on, why the obvious key is unsafe,
and what the shipping CLI cache already does.

\section{Why OID Alone Is Unsafe}

A summary cache keyed on \texttt{(volume\_uuid, oid)} would
deduplicate unchanged subtrees across runs. It would also return
stale summaries the first time the volume mutates a subtree the
cache already knows.

\textbf{Observation.} The FS-tree root's logical OID stays stable
across mutations; every other identity field changes. EX-06 mutated
the proof fixture, recorded the FS-tree root identity before and
after, and observed:

\begin{center}
\begin{tabular}{ll}
\toprule
field & across mutation \\
\midrule
logical OID                & stable \\
physical address (paddr)   & changed \\
object XID                 & changed \\
object checksum            & changed \\
block hash                 & changed \\
\bottomrule
\end{tabular}
\end{center}

A cache keyed on OID would return the pre-mutation summary for the
post-mutation tree, and the user would see a treemap that does not
match the disk. The OID names \emph{some} version of the object;
it does not name particular content. A cache that wants content
identity must use a content identifier.

\section{The Shape Of A Safe Cache Key}

A safe cache key distinguishes entries that share an OID but
differ in content. The minimum sufficient key is a tuple from
several layers:

\begin{description}
  \item[Source identity.] A volume UUID, a container UUID, and an
    image fingerprint (for detached sources). Distinguishes one
    APFS volume from another, and distinguishes the same volume's
    different image dumps.
  \item[Parser version.] A monotonic identifier for the body decoder
    and the namespace builder. A parser change invalidates every
    summary, because a summary's interpretation depends on the
    parser that produced it.
  \item[Summary schema version.] A monotonic identifier for the
    summary structure itself. A schema change is independent of a
    parser change.
  \item[OMAP context.] Whether the OID was resolved through the
    container OMAP or the volume OMAP. A collision between the two
    name spaces is rare but representable.
  \item[OID.] The virtual identifier.
  \item[Object XID.] The version of the object the summary was built
    for.
  \item[Physical address.] Useful as a cross-check; an OMAP that
    resolves \texttt{(oid, xid)} to a different paddr than the cache
    expects invalidates the entry.
  \item[Checksum or content hash.] The strongest individual signal.
    A block whose checksum has changed must not reuse a summary
    keyed on the old checksum.
  \item[Expected type and subtype.] Defensive. An OID that has been
    re-allocated to a different object class is a new object.
\end{description}

The key is not a small integer: nine fields, eight of which the
parser already computes during a read, the ninth (parser version)
constant per build.

\section{What EX-07 Proved}

EX-07 asked a narrow question: under exact node-identity matches on
every field of the key above, does subtree reuse ever produce a
wrong summary?

\textbf{Observation.} EX-07 found zero false reuse for exact
node-identity matches on the detached lab corpus. Two parses that
see the same node identity get the same summary; reusing one for
the other is safe.

The result is not a production cache theorem. EX-07 ran one corpus,
on detached images, with a small mutation set. A production cache
that consumes the raw parser must also resolve cache storage with
crash-safe writes (a corrupting cache produces
stale-and-wrong instead of absent-and-correct) and an explicit
invalidation policy covering parser-version mismatch, source
identity mismatch, feature drift, checkpoint rollback, snapshot
switch, unknown record-body types, and any new metric absent from
the pre-existing schema. The Real metric of chapter~8 is the most
recent example of a column a pre-existing cache would not have
recorded.

\section{The Shipping CLI Cache (EX-30)}

The CLI ships a v1 cache behind the \texttt{--cache} flag (off by
default). It operates on the fallback walker, not the raw parser:
the safe-reuse work above remains future for raw-mode subtree
summaries. The v1 design is intentionally smaller than what EX-07
gestures at.

\textbf{Mechanism.} On a cache miss, the full scan runs, the
output is written to a content-addressed entry under
\texttt{~/Library/Caches/apfs-fastindex/}, and the entry is keyed
on a probe-derived signature of the scanned directory tree. On a
hit, the cached entry is replayed without walking the tree.

\textbf{Observation (EX-30).} The cache works on stable trees and
does not on actively-churning ones:

\begin{center}
\begin{tabular}{lrrl}
\toprule
target & first scan & rescan (\texttt{--cache}) & speedup \\
\midrule
project \texttt{docs}     &  0.33~s &  0.007~s & 47$\times$ \\
project root              &  0.59~s &  0.12~s  & 4.9$\times$ \\
\texttt{/Users/kai}       & 46.6~s  & 45.4~s   & 1.0$\times$ \\
\bottomrule
\end{tabular}
\end{center}

Two factors compound on \texttt{/Users/kai}-class trees. First, the
probe uses naive \texttt{read\_dir} plus \texttt{symlink\_metadata}
per child; the walker uses \texttt{getattrlistbulk}, which batches
metadata reads. On 172k directories the probe pays roughly 37~s
versus the walker's roughly 9~s --- a cache hit costs more than no
cache. Second, dozens of subdirectories under \texttt{/Users/kai}
mutate continuously (browser caches, Spotlight, log rotation, Xcode
derived data); the signature almost never matches the cached
signature, every cache write is wasted.

For stable trees (project directories, \texttt{/Applications},
read-only system paths) the v1 cache delivers a 5--50$\times$
speedup. The CLI prints \texttt{cache: hit} to stderr on hits; the
documentation labels the flag ``useful for stable trees; not
recommended for active home directories.''

\section{What R4 v2 Needs}

The v1 cache becomes useful on home-directory-scale trees when two
pieces land. Both belong to future work; neither is in the shipping
binary.

\begin{enumerate}
  \item \textbf{Walker-emitted signature.} The parallel walker
    accumulates each directory's signature inline, while it is
    already calling \texttt{getattrlistbulk}. The separate
    directory-only probe disappears; the cache-hit cost drops from
    roughly 37~s to roughly 0~s.
  \item \textbf{Per-subtree caching.} One signature per directory
    rather than per scan. On rescan the walker descends
    incrementally: when a directory's signature matches the
    cached one, the subtree's entries and aggregates are reused
    without descent. A few churning subdirectories invalidate
    themselves; the unchanged 99\,\% stays cached.
\end{enumerate}

Together these turn a 45-second rescan on \texttt{/Users/kai} into
roughly one second when 10\,\% of subdirectories have changed.
That is the R4 v2 promise; an EX-* in the next phase will validate
it against a forced full rescan.

\section{What This Chapter Forecloses}

\paragraph{No bare-OID cache.} The parser does not key reuse on
OID alone, even with parser-version qualification. The mutation
behaviour above makes a bare-OID cache unsafe in principle.

\paragraph{No reuse across mismatched scan XIDs.} A summary
computed at scan XID $a$ is not reused at scan XID $b$. Object
identity includes the XID; relaxing that turns the cache into a
stale-state producer. The cache may compare keys across runs only
when the keys agree on every field, scan XID included.

EX-06's discipline --- record every plausibly-load-bearing identity
field at every relevant boundary before deciding which ones matter
--- generalises beyond caching. EX-12's OMAP probe used the same
habit, recording every resolved \texttt{(oid, paddr, xid)} sample
before deciding which mattered. The cost of recording is small; the
cost of discovering later that a key field was never captured is
large.
