\chapter{Support, Readability, and the Fallback}

A source the product makes a correctness claim on is
\emph{supported}. A source that can merely be opened and decoded
is \emph{readable}. This chapter develops the vocabulary, the
current allowlist, and the fallback path the product runs when the
boundary refuses a source.

\section{Four Properties}

Four properties characterise a source's relationship to the parser.
The order matters; each builds on the last.

\begin{description}
  \item[Readable.] Raw bytes can be opened and sampled. A device file
    exists and \texttt{read()} returns data.
  \item[Parsable.] Checkpoint and root discovery succeed; raw
    traversal runs without aborting.
  \item[Validatable.] The parser's output can be compared to a named
    oracle for the requested mode. A namespace can be diffed
    against a POSIX traversal; a logical size against
    \texttt{st\_size}.
  \item[Supported.] The source is in the product allowlist for the
    requested mode. Validation has been performed; results were
    correct; the source class belongs in production.
\end{description}

The first three are technical. The fourth is editorial: the
project's decision about which sources are stable enough to make a
correctness claim on. A source can be readable, parsable, and
validatable in one moment and still not be supported because the
moment is not durable.

A mounted live volume is readable. It may be parsable on a still
moment. It may be validatable against a momentary POSIX snapshot.
It is not supported, because the validation is not reproducible:
the next moment moves the visible checkpoint, the next POSIX
snapshot disagrees with the parser snapshot, and no row can be
attributed to one moment. The general rule: support a source class
only when the validation recipe is reproducible.

\section{The Current Allowlist}

The raw-mode allowlist is intentionally short.

\begin{itemize}
  \item Detached, unencrypted, image-backed APFS containers. The
    \texttt{.dmg} layer fixes the visible state at detach time.
  \item Caller-supplied raw APFS container devices under
    \texttt{/dev/disk*} or \texttt{/dev/rdisk*} where the caller
    has committed that the checkpoint will not advance.
  \item Explicitly stable APFS sources --- a tightly controlled lab
    volume where the operator pins the state.
\end{itemize}

What is not on the list: live mounted startup disks, FileVault
volumes the user is using, multi-device or Fusion containers,
sealed system volumes, snapshots opened through the operating
system's snapshot framework, and any container whose
\texttt{omap\_phys.flags} indicate an encryption operation in
flight.

The allowlist will grow only after a paired oracle and a new probe
clear the corresponding entry in the support matrix.

\section{Live-Volume Raw Mode (EX-28)}
\label{sec:ex28-live-raw}

A live mounted volume's \texttt{/dev/diskNsM} device node is not
on the allowlist: a live mounted volume's metadata advances during
the scan window and SR-014 checkpoint selection has no way to pin
the visible state.

EX-28 asks the narrower question: with root privileges, does the
existing raw parser produce a self-consistent shape on
\texttt{/dev/diskNsM}, and does that shape agree with what the
fallback walker emits from the same mount point in the same scan
window?

The harness lands as a Rust integration test in
\texttt{crates/apfs-fastindex/tests/ex28\_live\_parity.rs},
gated on the environment variables
\texttt{APFS\_FASTINDEX\_EX28\_LIVE\_DEVICE} and
\texttt{APFS\_FASTINDEX\_EX28\_MOUNT\_POINT}. Without those, the
test is a clean no-op and \texttt{cargo test --release} skips it.
With them set and the test run as root, the harness:

\begin{enumerate}
  \item Tries three successive raw scans of the device. Each
    must publish a \texttt{selected\_checkpoint} (no SR-014
    fail-closure under live concurrent writes), with pairwise
    symmetric difference of their \texttt{NamespaceEntry} sets
    $\leq 200$ paths.
  \item Runs one fallback-walker scan of the mount point and
    compares it against the raw scan; symmetric difference must
    be $\leq 1000$ paths to absorb the
    raw-can-read-paths-the-fallback-cannot asymmetry (TCC, sealed
    system volume reads).
\end{enumerate}

The comparator behind both checks is
\texttt{apfs\_fastindex::parity::compare\_namespace\_shapes},
which returns a structured \texttt{ShapeDiff} with
\texttt{only\_in\_left}, \texttt{only\_in\_right}, and per-field
\texttt{mismatches}. It is exposed as a public crate surface so
any future tooling can use it.

\textbf{Observation (2026-05-20).} On the project owner's Apple
silicon host, macOS refused to open \texttt{/dev/disk3s1} for raw
reads, returning \texttt{EPERM} on the first \texttt{read(2)} of
block zero even under \texttt{sudo}. The harness classifies this
as \texttt{LiveScanOutcome::BlockedByKernel} and records the
\texttt{live\_raw\_blocked\_by\_kernel} verdict. The kernel's
storage-system security policy refuses raw block access to the
live boot data partition independently of POSIX permissions; SIP
and the sealed-system-volume seal both contribute. Neither the
successive-scan stability nor the raw-vs-fallback parity
assertions could be evaluated because the raw backend never
opened.

\textbf{Product implication.} The ``Scan as Administrator''
shell action does not graduate to raw mode on a stock Apple
silicon macOS host. The privileged-subprocess shape still serves
a purpose --- unlocking TCC-restricted user-data paths the
fallback walker hits \texttt{EACCES} on --- but it stays on the
fallback walker even under root. The raw fast path is unchanged
in its original scope (detached \texttt{.dmg} images and
caller-pinned raw devices on a quiesced volume). EX-28's
contribution to the codebase is the reusable shape-parity
comparator (\texttt{apfs\_fastindex::parity}) and the env-gated
integration harness, both rerunnable on future hosts or future
macOS versions without code changes.

The live-disk row of the support matrix: \emph{not supported on
stock macOS; raw read returns \texttt{EPERM} under \texttt{sudo};
the fallback walker remains authoritative, including the
privileged-subprocess invocation that unlocks TCC-restricted
user-data paths.}

\section{Local-Snapshot Enumeration (EX-29)}
\label{sec:ex29-snapshots}

The product surfaces local snapshots as a status-bar row because
their bytes are part of \texttt{df}'s ``used'' total and the
volume's Finder-reported capacity, but the scanner walks only the
live volume's fs-tree; without a snapshot row, those bytes look
missing.

EX-29's deliverable is enumeration: a count of user-visible local
snapshots, with names. Reclaimable bytes are out of scope on stock
Apple silicon. There is no public read-only oracle for per-snapshot
reclaimable bytes (\texttt{tmutil thinlocalsnapshots} reports the
number, but it deletes the snapshot in the process), and EX-28's
verdict rules out the raw-extent-set-diff path.

Enumeration is unprivileged. The \texttt{snapshots} module shells
out to \texttt{tmutil listlocalsnapshots <mount>} and
\texttt{diskutil apfs listSnapshots <mount>}, parses each output,
and classifies. SR-020's sealed-system filter excludes
\texttt{com.apple.os.update-*} (the boot snapshot the user cannot
delete).

\textbf{Observation (2026-05-20).} The project owner's host has
zero user-visible TM local snapshots and one sealed-system
OS-update snapshot on \texttt{disk3s1s1}. The classifier returned
\texttt{NoUserSnapshots\{sealed\_system\_excluded\_count = 1\}}.
EX-23 found the same shape five days earlier; on a single-user
development host without Time Machine configured this is the
expected steady state.

The integration harness in
\texttt{tests/ex29\_snapshot\_enumeration.rs} mirrors EX-28's
pattern: an unconditional unprivileged test runs the classifier;
a gated test exercises a raw extent diff between a mounted
snapshot's device node and the live volume via
\texttt{parity::compare\_namespace\_shapes}. The gated test reuses
EX-28's \texttt{LiveScanOutcome} classifier, so an \texttt{EPERM}
on the snapshot device is recorded rather than failing the test.

Support matrix row: \emph{enumeration validated; reclaimable bytes
unclaimed.}

\section{Fail-Closed Triggers}

A source outside the allowlist fails closed before any row is
emitted. The categorical triggers, with the chapter that develops
each:

\begin{description}
  \item Ambiguous or malformed checkpoint state (chapter~3).
  \item Non-contiguous checkpoint descriptor layout (chapter~3).
  \item Unsupported incompatible feature bits in the container or
    volume (chapter~3, chapter~5).
  \item OMAP-open or OMAP-lookup encryption flags (chapter~4).
  \item Unknown bits in any \texttt{flags} field the parser inspects
    (chapter~4, chapter~7).
  \item Encryption keys or hardware-backed key requirements the
    parser cannot reach (chapter~4).
  \item Fusion or multi-device storage requirements (chapter~3).
  \item Requested merged-root or boot-root semantics (chapter~13).
  \item Snapshot or sealed-volume semantics outside the target mode
    (chapter~13).
  \item No validated corpus for the source class.
\end{description}

Each trigger is a typed error with a substring naming the rule.
The caller dispatches on the error rather than retrying.

\section{The Fallback}

When the parser refuses a source, the fallback path emits the
same row shape through supported operating-system APIs.

\textbf{Observation.} A POSIX-traversal fallback in both Python
(\texttt{fallback\_traversal.py}) and Rust (\texttt{fallback.rs})
emits \texttt{NamespaceEntry} and \texttt{DirectoryAggregate} rows
that, on the proof fixture, match the raw-mode output of the same
volume exactly. Same entry count, same paths, same
\texttt{entry\_kind}, same \texttt{logical\_size}, same
\texttt{symlink\_target}, same per-directory aggregates.

The fallback walker uses \texttt{os.walk} plus \texttt{os.lstat} plus
\texttt{os.readlink} in the Python version, and the equivalent
\texttt{read\_dir} plus \texttt{symlink\_metadata} plus
\texttt{read\_link} in the Rust version. Both paths apply the SR-017
precedence rule at the trivial step: \texttt{st\_size} for regular
files, \texttt{readlink} byte length for symlinks. Both apply the
unique-inode aggregate policy from chapter~8.

The row contract is mode-independent: the caller does not need
to know which backend produced the row. The field set is
identical.

The fallback is not slow. Chapter~12 records 313~k entries per
second on warm directory trees with four workers
(2.47$\times$ the post-micro-optimisation single-threaded
throughput, 1.82$\times$ the pre-perf baseline). On a cold
whole-machine scan the disk is the bottleneck, not the parser.

\section{Resilience In The Fallback}

A whole-machine scan as a normal user encounters paths that the
user does not have permission to read: sandboxed application data,
sealed-system directories, daemon-owned caches. The fallback walker
must continue across these without aborting and without producing
silently incomplete totals.

\textbf{Observation.} Per-entry \texttt{EACCES}, \texttt{EPERM}, and
\texttt{ENOENT} are recorded in \texttt{parser\_output.walk\_skips}
with a reason; the walk continues. The same log carries the
\texttt{firmlink\_overlay\_source} and \texttt{duplicate\_dir\_inode}
skip kinds introduced by the dedup pass below. A normal-user
\texttt{/} scan on the host the project tested completes with
roughly five hundred recorded skips (the exact count varies with
the host's app sandbox profile and the firmlink set) and reports
them in the output document. The caller can render the skip list
as a diagnostic without it affecting the row stream.

The walker also skips top-level system directories that exist only as
APFS-internal artefacts on the boot disk
(\texttt{.fseventsd}, the system snapshot mountpoints) by name,
rather than letting the operating system raise an error on them.
That skip is documented in the source so it can be revisited as
support broadens.

\section{Parallel Fallback}

\textbf{Observation.} The fallback walker accepts a
\texttt{--threads N} flag. The default is
\texttt{min(hw.physicalcpu, 4)}; the lower bound is 1, the upper
bound clamped to 4. Per-worker \texttt{BulkReader} owns its own
64~KiB \texttt{getattrlistbulk} buffer; a shared work queue
distributes directories; per-worker entry vectors are
concatenated and re-sorted on the main thread.

The default's ceiling at 4 is not arbitrary. The single APFS
container holds a B-tree contention point that pre-Mojave was
described as a global mutex (Szorc 2018) and that the Apple
Developer Forums position (Apple DTS, 2025) describes as having
been largely addressed in macOS~10.14 while leaving a residual
per-container cost. Empirical measurement on Apple silicon
(EX-25, chapter~12) shows the throughput envelope peaking at
$T = 4$ with 2.47$\times$ speedup over single-threaded, then
\emph{regressing} at $T = 8$ and falling further at $T = 14$. At
$T = 8$ the total system CPU is $4\times$ the $T = 1$ value while
the throughput is only $1.94\times$; at $T = 14$ system CPU is
$9.3\times$ for $1.38\times$ throughput --- the kernel pays
super-linearly more work to deliver sub-linearly more answers.
The 4-thread ceiling stays clear of that regime by construction.

The parallel walker preserves the serial walker's contract by
shape, asserted by a synthetic-tree regression test
(\texttt{parallel\_walker\_matches\_serial\_shape}): identical
entries, aggregates, and skip lists; the only difference is the
order in which directories are visited, which the final sort
normalises.

Live progress events flow through a dedicated sampling thread
inside the parallel scheduler at a 250~ms cadence (matching the
serial path), so the CLI's \texttt{-{}-progress} stream and any
user-space consumer of it can tick smoothly on a whole-machine
scan that is otherwise silent for tens of seconds. The in-process
FFI consumed by the native shell (chapter~13) exposes the same
stream through \texttt{apfs\_scan\_directory\_with\_progress},
whose callback the sampling thread invokes from the library side.
The SwiftUI shell hands in a closure that updates an
\texttt{@Published} \texttt{ProgressSnapshot} on the main actor;
the determinate progress island follows it. One source of truth,
two delivery surfaces.

\section{Firmlink Overlay Dedup}

\textbf{Observation.} On macOS the user-visible startup root is a
join of the System volume and the Data volume. Specific paths ---
\texttt{/Users}, \texttt{/Applications}, \texttt{/Library},
\texttt{/private}, \texttt{/usr/local}, and a few others --- are
exposed both at their canonical location \emph{and} at
\texttt{/System/Volumes/Data/<source>} via firmlinks. The same
underlying inode is visible through two paths. A naive recursive
walker descends both and double-counts everything inside the
overlay.

The fallback walker, when the root is \texttt{/}, reads
\texttt{/usr/share/firmlinks} at scan-start and builds a
\texttt{HashSet} of the \texttt{/System/Volumes/Data/<source>}
overlay paths. A descent into any such path is refused with a
\texttt{firmlink\_overlay\_source} skip; the canonical path
remains the one that contributes to the namespace and aggregates.

A second layer of safety, a per-scan
\texttt{Mutex<HashSet<(dev, ino)>>}, catches any directory
inode reached through a path the firmlink overlay set did not
cover. A second observation of the same inode through a different
path is recorded as a \texttt{duplicate\_dir\_inode} skip and the
walker does not recurse.

The effect on whole-machine scans is large. Before the dedup, a
\texttt{/} scan on the test host reported 5,250,000 entries;
after, 3,060,000 entries with seventeen
\texttt{firmlink\_overlay\_source} skips (one per existing
firmlink target) and zero \texttt{duplicate\_dir\_inode} skips
(the overlay layer caught everything). The forty-two per cent
that disappeared were not real files: they were the same files
counted twice through two paths.

\section{Auto-Dispatch}

The CLI binary \texttt{apfs-fastindex-scan} picks the backend
automatically based on the source path's shape.

\begin{itemize}
  \item A \texttt{.dmg} path or a \texttt{/dev/disk*} /
    \texttt{/dev/rdisk*} device path is dispatched to raw mode.
  \item A directory path is dispatched to fallback mode.
  \item The caller can override with \texttt{--mode raw},
    \texttt{--mode fallback}, or \texttt{--mode auto}.
\end{itemize}

Auto-dispatch makes the support boundary usable as a product
surface. The binary handles the dispatch and reports through
\texttt{correctness\_claim} and \texttt{not\_claimed} which mode
ran.

\section{Why The Boundary Is Permanent}

As the allowlist grows (chapter~13 tracks the open candidates),
each new source class gets its own line in the support matrix:
own allowlist row, own oracle, own probe. The boundary moves only
when there is evidence to move it.
