\chapter{Engineering Surface and Open Questions}

The previous chapters develop the manual's defended interior: one
volume, one pinned state, namespace, logical, allocated, and
clone-deduplicated size, fallback for sources outside the
allowlist. This chapter records the engineering surface around that
interior and the questions that remain open.

\section{The Native Shell}

The product surface is a SwiftUI app whose visualisation is laid
out by Rust and drawn by Core Graphics. The scanner is not a
subprocess; the Rust crate is linked into the app process as a
static library, and every call from Swift into the indexer crosses
a \texttt{cbindgen}-generated C ABI. There is no JSON
serialisation, no temp file, no WebKit, no JavaScript.

\subsection{The FFI Boundary}

The crate's C ABI is a small surface of opaque-handle
constructors, accessors, and destructors:

\begin{description}
  \item[\texttt{apfs\_scan\_directory(\_with\_progress)}]
    returns an \texttt{ApfsScan*}. The handle owns the
    \texttt{FallbackScanOutput} (entries, walk-skips,
    \texttt{correctness\_claim}, source descriptor), the indexed
    tree built from it, and a lazy path-string cache.
  \item[\texttt{apfs\_layout\_new(scan, subtree, depth, metric,
    dims)}] returns an \texttt{ApfsLayout*} whose flat
    \texttt{ApfsCell} array Swift reads as
    \texttt{UnsafeBufferPointer<ApfsCell>}: no per-cell call into
    Rust during paint, no array copy across the language boundary.
  \item[\texttt{apfs\_scan\_ext\_summary\_new}] returns an
    \texttt{ApfsExtSummary*} with the per-extension aggregate the
    right-hand panel reads.
  \item[\texttt{apfs\_scan\_search\_names}] returns the
    node-indices of substring matches sorted by the active metric.
\end{description}

Bulk data crosses the boundary as \texttt{(*const Item, usize)}
pairs Swift consumes via \texttt{UnsafeBufferPointer<Item>} ---
no copy, no per-element call into Rust. Per-handle
\texttt{\_free} functions release every allocation \texttt{\_new}
produced. The header regenerates from \texttt{src/ffi.rs} on every
change via \texttt{cbindgen}, so the C declarations cannot drift
from the Rust \texttt{\#[no\_mangle] extern "C"} surface.

\subsection{Division Of Labour}

\textbf{Rust owns:} the fallback walker, the firmlink dedup, the
indexed tree (one \texttt{TreeNode} per directory and per leaf,
parent and children indices, pre-computed
\texttt{value\_logical} / \texttt{value\_allocated} /
\texttt{value\_real} / \texttt{item\_count} aggregates), the
squarified treemap layout (a direct port of
\texttt{d3-hierarchy}'s \texttt{treemapSquarify}), the
$64 \times 64$ hit-grid, the per-extension aggregator behind the
right-hand panel, the substring-search index, and the colour
palette that decides a leaf's \texttt{fill\_rgb}.

\textbf{Swift owns:} the SwiftUI shell (toolbar, breadcrumb,
folder-tree side panel, ext-list side panel, search field,
status bar, depth stepper, metric picker), the AppKit treemap
view (one \texttt{NSView} whose \texttt{draw(\_:)} walks the cell
array grouped by colour to amortise \texttt{setFillColor}
calls), the right-click menu plumbing onto \texttt{NSWorkspace}
(\texttt{Open}, \texttt{Reveal in Finder}, \texttt{Copy Path},
\texttt{Move to Trash}), the sticky admin-mode flag and
privileged-subprocess invocation, the snapshot enumeration
status pill, and the Sparkle updater.

\subsection{Treemap Layout}

The layout is the standard squarified treemap of Bruls, Huijzen,
and van Wijk (2000), in the variant \texttt{d3-hierarchy}
implements. Children are sorted descending by the active metric;
greedy row-packing on the shorter side of the current rect adds
children to a row until the worst-case aspect ratio would worsen,
then lays the row out by dice (horizontal) or slice (vertical)
and recurses on the remaining strip. The Rust port preserves the
d3 expressions for $\alpha$ and $\beta$; for the same input
\texttt{(node\_index, value)} array and viewport the two
implementations produce the same rectangles.

A $64 \times 64$ axis-aligned spatial hash sits over the cell
array. Each cell is inserted into every bucket it overlaps; each
bucket is sorted depth-descending so the first containing match a
hit-test walk sees is the deepest cell at that point. Mouse-move
becomes a few-cell comparison rather than a walk through the
hundred-thousand-cell layout for a \texttt{/}-scan view.

When the metric picker is on Allocated (or Real) and a leaf's
\texttt{value\_allocated} (or \texttt{value\_real}) is
\texttt{None}, the cell's \texttt{fill\_rgb} is the unclaimed
grey rather than a colour from the extension palette. The
fail-closed register surfaces on the picture.

\subsection{Substring Search (EX-33)}
\label{ssec:search-perf}

The shell ships a substring search over already-enumerated names.
The C ABI exposes \texttt{apfs\_scan\_search\_names(scan,
needle\_ptr, needle\_len, limit)}, which returns a sorted slice
of matching node indices. The implementation is the result of a
two-round perf study.

\textbf{Round 1 (pre-EX-33).} The naive
\texttt{Vec<String>}-of-pre-lowercased-names plus
\texttt{str::contains} brought per-keystroke latency down from
multi-second to roughly 30--55~ms on the 1.56~M-entry
\texttt{/Users/kai} scan --- a 100$\times$ win, still 10--20$\times$
off the memory-bandwidth floor for 50~MB of name bytes.

\textbf{Round 2 (EX-33).} Two compounding costs remained: the
\texttt{Vec<String>} layout put each name behind its own heap
allocation (1.5~M random-pointer reads are 10--20$\times$ slower
than sequential bytes); and \texttt{str::contains} built its
searcher state once per call (1.5~M searcher constructions for
the same needle). The fix replaced the storage with a flat
\texttt{Vec<u8>} plus a cumulative \texttt{Vec<u32>} offset
table, and built one \texttt{memchr::memmem::Finder} per search
reused across every haystack.

\begin{center}
\begin{tabular}{lrrr}
\toprule
query                              & before & after & speedup \\
\midrule
\texttt{".txt"}                    & 32.5~ms & 10.1~ms & 3.21$\times$ \\
\texttt{"Library"}                 & 41.7~ms & 11.8~ms & 3.53$\times$ \\
\texttt{"node\_modules"}           & 53.2~ms &  9.5~ms & 5.59$\times$ \\
\texttt{"com.apple.developer"}     & 34.6~ms &  5.3~ms & 6.53$\times$ \\
zero-match (memory-bandwidth floor)& 27.5~ms &  2.3~ms & 12.0$\times$ \\
\bottomrule
\end{tabular}
\end{center}

Typical 4--12-character queries now median 9--12~ms on the 1.5~M
entry tree --- comfortably inside the ``feels live'' threshold
(100~ms) and within range of 60-Hz frame-rate (16~ms) for
hot-cache repeats. The shorter-needle high-match cases see modest
gains because their cost is dominated by ancestor walking
on millions of matches; long-needle queries hit the
memory-bandwidth floor.

The search is substring only; it does not perform an APFS-hash
lookup (chapter~9) and inherits no risk from the missing hash.

\subsection{What Is Not In The Shell}

\begin{itemize}
  \item Snapshot-content traversal: the SR-020/EX-23 oracle
    requires the privileged \texttt{mount\_apfs -s} step; EX-29
    confirmed reclaimable-bytes accounting is blocked by EX-28
    on the same host class.
  \item Live raw mode: EX-28 records the
    \texttt{live\_raw\_blocked\_by\_kernel} verdict. The
    privileged-subprocess shape still runs the fallback walker
    to unlock TCC-restricted paths, but does not graduate to
    raw mode.
  \item Code signing and notarisation.
\end{itemize}

\section{Boot-Root Semantics (EX-31)}

Gate~3 of the roadmap asked: define the semantic target for a
Finder-visible scan, build an oracle, compare raw System and Data
outputs against the user-visible tree, decide whether the mode is
API-only or raw-capable.

EX-28 foreclosed the raw comparison (the raw read returns
\texttt{EPERM} on the live boot data partition). EX-31 fills in
the rest: the only product mode the app can deliver on a typical
macOS install is Finder-visible startup root via POSIX traversal,
and that is exactly what the existing fallback walker already
does. POSIX traversal transparently follows firmlinks; the user
scanning \texttt{/} gets one tree matching what Finder shows ---
\texttt{/Users} and \texttt{/System/Library} both appear in
expected positions even though \texttt{/Users} lives on the data
volume and \texttt{/System/Library} on the sealed-system
snapshot.

\textbf{Observation.} On the project owner's host every
top-level path under \texttt{/} reports the same \texttt{st\_dev}
from \texttt{stat()} / \texttt{fstatat()} --- the kernel exposes
the system volume's dev for every path under \texttt{/} regardless
of which volume the bytes physically live on. The
\texttt{st\_dev}-based ``is this a different filesystem?''
heuristic carries no firmlink information; the firmlink table at
\texttt{/usr/share/firmlinks} is the only authoritative source.
The walker reads it at scan-start for the dedup pass of
chapter~11; nothing else is needed.

Verdict: \texttt{api\_only\_already\_correct}. A future polish
would surface a more specific mode label
(\texttt{fallback\_finder\_visible} rather than bare
\texttt{fallback}) so consumers can distinguish a merged-Finder
scan from a single-external-volume scan; the current envelope
does not. Slotted for a follow-up.

\section{Encryption (EX-32 Scratchpad)}

FileVault and per-file encryption sit at the most opaque corner
of APFS. The on-disk structures are named in the reference
(\texttt{apfs\_keybag\_loc}, \texttt{apfs\_kb\_locker\_t},
\texttt{apfs\_keybag\_entry\_t}, \texttt{crypto\_state\_t},
\texttt{wrapped\_meta\_crypto\_state\_t}, \texttt{cp\_key\_class\_t},
\texttt{INCOMPAT\_ENC\_ROLLED}), but field-level semantics, the
cryptographic primitives at each layer, and the key-derivation
chain are described only at a behavioural level. Bit layouts,
KDF input strings, per-block IV derivation, and the Secure-Enclave
relationship come from third-party reverse-engineering
(\texttt{apfs-fuse}, \texttt{linux-apfs-rw}, FileVault-to-APFS
transition writeups).

EX-32 is a research scratchpad, not a feature. The current crate
exposes \texttt{crypto\_meta} as a non-production module whose
public symbols decode keybag and per-file crypto-state shapes
against synthetic fixtures. The deliverable is documentation; no
decryption path is implemented and none is planned without an
oracle. The encryption row of the support matrix stays at
\emph{hard-stop at OMAP open}.

\section{Open Questions}

\subsection{Exclusive and Shared Bytes}

The Real metric of chapter~8 charges each inode its share of
clone-deduplicated bytes via the extent-reference tree.
\emph{Exclusive} bytes --- bytes only this inode owns, after
accounting for clones \emph{and} snapshots --- additionally
requires the snapshot-metadata tree to be opened and joined
against per-inode extents. \emph{Shared} bytes are the
complement. Neither has been validated against a public oracle;
both are open.

\subsection{Snapshot-Assisted Scanning}

The mechanism for honest scanning of a live machine. The walker
already accepts a \texttt{-{}-snapshot <mountpoint>} flag that
runs the fallback against an already-mounted snapshot directory;
the missing piece is the shape-parity oracle.

\textbf{Privilege boundary (SR-020).}
\texttt{fs\_snapshot\_create(2)} requires root \emph{and} the
\texttt{com.apple.developer.vfs.snapshot} entitlement. The same
gate applies to delete, rename, mount, and revert.
\texttt{mount\_apfs -s} also requires root.
\texttt{tmutil localsnapshot} is the only unprivileged create
path, available only on Time-Machine-included volumes.

\textbf{Reproducer.} The probe is re-runnable; EX-23's clean dev
workstation found no user-visible TM snapshots and emitted
\texttt{blocked\_no\_snapshots\_at\_all}. The privileged rerun
reads:

\begin{minted}{bash}
tmutil localsnapshot /
tmutil listlocalsnapshots /
sudo mkdir -p /tmp/apfsfi-ex23-snap
sudo mount_apfs -s <snapshot-name> / /tmp/apfsfi-ex23-snap
python3 docs/research/experiments/EX-23-snapshot-shape-parity/artifacts/probe_ex23.py
sudo umount /tmp/apfsfi-ex23-snap
\end{minted}

A positive verdict would unlock the snapshot-mount cell of the
support matrix from \texttt{not\_claimed} to
\texttt{fallback\_supported}.

\subsection{Persistent Incremental Cache}

The R4 v1 cache (chapter~10) ships on stable trees and degenerates
on home-directory-scale churn. The R4 v2 promise --- walker-emitted
signatures plus per-subtree caching --- needs a probe against
the EX-30 baseline. Acceptance: 9.62~s warm baseline,
$\leq$1.0~s unchanged-tree rescan, linear interpolation under
partial mutation.

\subsection{Encrypted Surfaces}

FileVault, per-file class keys, hardware-tied key wrapping. The
parser fails closed on every encryption flag (chapter~4); opening
support requires deciding which key-access classes the reader is
willing to use. EX-32 is the documentation scratchpad; no
feature follows from it without an oracle.

\subsection{Live Boot Disk Raw Mode}

EX-28's verdict
\texttt{live\_raw\_blocked\_by\_kernel} stands on stock Apple
silicon under SIP. The harness remains re-runnable on future
hosts or future macOS versions with different security policies
without code changes.
