\chapter*{Glossary}
\addcontentsline{toc}{chapter}{Glossary}

The vocabulary the manual uses, in alphabetical order. Terms that
appear in code identifiers carry the typewriter font of the
identifier as well as the prose definition.

\begin{description}
  \item[Allocated size.] Bytes the filesystem has assigned to a
    file's content. Distinct from logical size (the user-visible
    byte length) and from exclusive size (bytes only this inode
    owns). R2-A: landed for ordinary, clone, hard-link, symlink,
    and directory case classes; fail-closed for sparse files and
    decmpfs-compressed files. Emitted as
    \texttt{allocated\_size: Option<u64>}; \texttt{None} indicates
    a deliberately-not-claimed row.
  \item[\texttt{alloced\_size}.] The
    \texttt{j\_dstream\_t.alloced\_size} on-disk field. The
    source for the R2-A allocated-size column on the case classes
    that emit a value. On macOS-produced volumes the field equals
    \texttt{round\_up(ds\_size, blocksize)}; on sparse files this
    overstates the actual allocation by the sparse-bytes hint.
  \item[APSB.] APFS volume superblock
    (\texttt{apfs\_superblock\_t}). Reached through the container
    OMAP at the scan XID; carries the volume OMAP, the FS-tree
    root, feature masks, role, UUID, and statistics.
  \item[Block zero.] The fixed locator block at the start of an
    APFS container. Carries enough metadata to find the
    checkpoint descriptor area but is not itself authoritative for
    the live container state.
  \item[\texttt{ApfsCell}.] The flat, \texttt{\#[repr(C)]} record
    the Rust renderer hands across the FFI as a contiguous
    array. Holds the cell's rectangle, depth, the index of the
    tree node it represents, a flag bitset
    (\texttt{is\_dir}, \texttt{is\_symlink}, label-strip
    reserved), and a precomputed \texttt{fill\_rgb}. Read by
    Swift as an \texttt{UnsafeBufferPointer<ApfsCell>} with no
    array copy.
  \item[\texttt{ApfsLayout}.] The Rust handle that owns the
    laid-out \texttt{ApfsCell} array and the spatial-hash index
    used for hit-testing. Constructed by
    \texttt{apfs\_layout\_new}, freed by
    \texttt{apfs\_layout\_free}. One \texttt{ApfsLayout}
    represents one \texttt{(subtree-root, depth, metric,
    width$\times$height)} configuration.
  \item[\texttt{ApfsScan}.] The Rust opaque handle the C ABI
    returns from \texttt{apfs\_scan\_directory}. Owns the
    \texttt{FallbackScanOutput} (entries, walk-skips,
    \texttt{correctness\_claim}, source descriptor) and the
    indexed tree built from it. Released by
    \texttt{apfs\_scan\_free}.
  \item[\texttt{BulkReader}.] The Rust struct that owns a
    reusable 64~KiB \texttt{getattrlistbulk} buffer and one output
    \texttt{Vec<BulkEntry>} across the whole walk. The
    micro-optimisation pass introduced it to replace the
    per-directory fresh allocation.
  \item[\texttt{cbindgen}.] The Rust crate that walks
    \texttt{src/ffi.rs} and emits the C header Swift imports. Run
    by the \texttt{build.rs} on every change so the C declarations
    cannot drift from the Rust \texttt{\#[no\_mangle] extern "C"}
    surface. The generated header lives in
    \texttt{target/<profile>/include/apfs\_fastindex.h}.
  \item[Checkpoint descriptor area.] A contiguous run of blocks
    that hold the rotating ring of container superblocks and
    their associated checkpoint maps. The authoritative live NXSB
    lives here.
  \item[Checkpoint map.] A block carrying ephemeral-OID mappings
    for the selected checkpoint. Walked in order from
    \texttt{nx\_xp\_desc\_index}; the last block carries the
    \texttt{CHECKPOINT\_MAP\_LAST} flag.
  \item[decmpfs.] The Apple compression xattr family. Small
    compressed files store their data inline in
    \texttt{com.apple.decmpfs} with no \texttt{j\_dstream\_t};
    larger files use \texttt{com.apple.ResourceFork}. v1
    fails closed on the allocated-size column for any inode whose
    only size source is decmpfs, because no public reader has
    been oracle-validated against the macOS
    \texttt{st\_blocks * 512} for the resource-fork path.
  \item[Directory aggregate.] A per-directory total expressed
    under the unique-inode logical-sum policy: each inode
    contributes its logical size at most once per aggregate root,
    regardless of how many hard-linked paths inside that root
    name it. Allocated-size aggregates use the same rule plus a
    \texttt{None}-propagation step.
  \item[\texttt{DIR\_REC}.] Directory record. The FS-tree leaf
    that represents a directory entry; key carries the parent
    inode's file ID and the name; value carries the child file ID
    and the POSIX entry type.
  \item[Dstream.] An APFS data stream. Logical and allocated size
    plus a default-extent reference; carried inline in the inode's
    xfields as \texttt{INO\_EXT\_TYPE\_DSTREAM} or in a separate
    \texttt{j\_dstream\_id\_val\_t} record.
  \item[Ephemeral object.] An APFS object that lives in memory
    between transactions and is written into the checkpoint data
    area at commit. The checkpoint map identifies it by
    \texttt{(oid, paddr)}.
  \item[Fail closed.] The project's discipline of refusing
    unknown, malformed, or out-of-allowlist input with a typed
    reason, rather than producing a partial or best-effort
    result.
  \item[Fallback path.] The supported-API backend that runs when
    raw mode rejects a source. POSIX walk plus \texttt{lstat}
    plus optional \texttt{getattrlistbulk}; emits the same row
    shape as raw mode.
  \item[Firmlink.] A macOS presentation mechanism that joins
    selected paths from a System volume to corresponding paths on
    a Data volume to produce the user-visible startup namespace.
    Out of v1 scope for raw mode. The fallback walker reads
    \texttt{/usr/share/firmlinks} at scan-start and skips the
    \texttt{/System/Volumes/Data/<source>} overlay paths so the
    canonical path is the one that contributes to namespace and
    aggregates.
  \item[\texttt{FsRecordDump} / \texttt{FsRecordRow}.] The Rust
    crate's structured output for the FS-tree record-family dump
    and one decoded record respectively. Field-level equal to the
    independent Python decoder on the proof fixture.
  \item[FS tree.] The volume-level B-tree of variable-length
    records keyed by \texttt{(object\_id, type)}. Holds directory
    entries, inodes, xattrs, sibling-link records, dstream
    records, and extents.
  \item[\texttt{getattrlistbulk}.] macOS syscall that returns
    metadata for many directory entries per call. Used by the
    fallback walker for the bulk perf path. Adding inode-required
    attributes (size, mode, mtime) to the attribute mask used to
    be thought a load-bearing cost; EX-24 falsified that for
    modern Apple silicon.
  \item[Hit-grid.] A $64 \times 64$ axis-aligned spatial
    hash sitting over the \texttt{ApfsCell} array. Each cell is
    inserted into every bucket it overlaps; each bucket is then
    sorted depth-descending so the first containing match a
    hit-test walk sees is the deepest cell at that point.
    Constant-time lookup turns mouse-move into a few-cell
    comparison even on a \texttt{/}-scan layout.
  \item[Hard link.] A directory entry that names an inode already
    named by another directory entry. APFS represents the
    relationship through SIBLING\_LINK and SIBLING\_MAP records;
    the unique-inode aggregate policy ensures the inode's logical
    size is counted once per aggregate root.
  \item[Indexed tree.] The in-memory \texttt{Tree} the FFI
    builds once per scan from the flat
    \texttt{Vec<NamespaceEntry>}. One \texttt{TreeNode} per
    directory and per leaf, carrying parent and children indices,
    own logical and allocated sizes, and post-order-computed
    \texttt{value\_logical}, \texttt{value\_allocated}, and
    \texttt{item\_count} aggregates. Random-access reads by
    index are constant-time; consumers need the indexed shape
    for the breadcrumb, the tree-list panel, the squarify
    recursion, and the hit-test loop.
  \item[\texttt{INODE}.] Inode record. The FS-tree leaf that
    carries file/dir/symlink identity, mode, parent ID, link
    count, internal flags, and an xfield tail.
  \item[Logical size.] User-visible byte length. The v1 size
    metric, equal to POSIX \texttt{st\_size} for regular files
    and to the symlink target byte length for symlinks. Distinct
    from allocated, exclusive, shared, and snapshot-retained
    sizes.
  \item[Max XID.] The XID argument to an OMAP lookup. Returns the
    entry with the highest stored XID less than or equal to the
    requested value. The whole point of pinning the scan XID.
  \item[\texttt{NamespaceEntry}.] One row of the namespace
    output: path, entry kind, file ID, logical size, optional
    allocated size, optional symlink target. Stable across raw
    and fallback backends.
  \item[Native shell.] The SwiftUI macOS app whose visualisation
    is laid out by Rust and drawn by Core Graphics. The Rust
    crate is linked into the app process as a static library;
    every call from Swift into the indexer crosses a
    \texttt{cbindgen}-generated C ABI. No subprocess, no JSON
    serialisation, no WebKit. Chapter~13 documents the
    architecture; the repository still ships
    \texttt{viz/index.html} as a standalone
    drop-a-file-into-browser surface.
  \item[NX / NXSB.] APFS container superblock
    (\texttt{nx\_superblock\_t}). The block-zero copy is the
    locator; the authoritative selected NXSB lives in the
    descriptor area at the highest valid checkpoint XID.
  \item[OID.] Object identifier. A virtual address into an APFS
    OMAP's name space. Not by itself a safe cache key
    (chapter~10).
  \item[OMAP.] Object map. The B-tree that resolves virtual OIDs
    to physical addresses under a chosen scan XID. Container and
    volume each have their own.
  \item[Paddr.] Physical block address. The on-disk location of an
    APFS object. Returned by an OMAP lookup for virtual objects;
    equal to \texttt{o\_oid} for physical objects.
  \item[Parallel walker.] The default fallback scheduler.
    Per-worker \texttt{BulkReader}, shared work queue, per-worker
    entry vectors merged on the main thread. Default thread count
    \texttt{min(hw.physicalcpu, 4)}, clamped to \texttt{[1, 4]}.
    The ceiling stays clear of the APFS container-lock regime
    that fires at $T > 4$ on Apple silicon.
  \item[Parsable.] Source property: checkpoint and root discovery
    succeed; raw traversal runs. Necessary but not sufficient for
    ``supported.''
  \item[\texttt{-{}-progress}.] CLI flag emitting stderr
    progress events at a 250~ms cadence. The parallel walker
    delivers events through a dedicated sampling thread so the
    cadence matches the serial path.
  \item[Raw mode.] The native APFS-parsing backend. Used when the
    source is in the raw-mode allowlist (detached \texttt{.dmg},
    caller-supplied raw device, explicitly stable APFS source).
  \item[Readable.] Source property: raw bytes can be opened and
    sampled. Cheap. Not, on its own, evidence of correctness.
  \item[Scan XID.] The transaction identifier the parser pins for
    the duration of a run. All object resolution is performed
    against this XID. Picked as the highest checksum-valid NXSB
    XID in the descriptor area.
  \item[Sibling link / sibling map.] Hard-link records.
    \texttt{j\_sibling\_link\_val\_t} carries a hard-link's
    sibling-group identifier and the linked name;
    \texttt{j\_sibling\_map\_val\_t} closes the mapping from
    sibling ID to inode.
  \item[Snapshot, APFS.] A read-only view of a volume at a
    point in time, named in the volume's snapshot meta tree.
    Sealed-system snapshots (\texttt{com.apple.os.update-*})
    are out of v1 scope. R2-B's
    \texttt{-{}-snapshot <mountpoint>} flag accepts an
    already-mounted snapshot directory; the mount step itself is
    privileged.
  \item[Squarify.] The treemap-layout algorithm of Bruls,
    Huijzen, and van Wijk (2000), in the variant
    \texttt{d3-hierarchy}'s \texttt{treemapSquarify} implements.
    Greedy row-packing on the shorter side of each rect; rows
    grow until the worst-case aspect ratio would worsen, then
    lay out by dice (horizontal) or slice (vertical) and recurse
    on the remaining strip. The Rust port preserves the d3
    expressions for \texttt{alpha} and \texttt{beta}, so the
    two implementations produce the same rectangles for the
    same input.
  \item[Sparse bytes.] \texttt{INO\_EXT\_TYPE\_SPARSE\_BYTES}:
    the count of bytes the sparse-allocation layer has not
    committed to disk. An allocation hint, not a logical size.
    Treated as diagnostic in the logical-size precedence and as
    the fail-closed gate for the allocated-size rule on sparse
    files.
  \item[Supported.] Source property: the source is in the product
    allowlist for the requested mode. Validation has been
    performed; results were correct; the source class belongs in
    production.
  \item[\texttt{-{}-threads N}.] CLI flag selecting the parallel
    walker's worker count. Default
    \texttt{min(hw.physicalcpu, 4)}; \texttt{N = 1} forces serial
    mode; larger values are accepted but pay the APFS-container
    contention cost documented in chapter~12.
  \item[Validatable.] Source property: parser output can be
    compared against a named oracle for the requested mode. A
    prerequisite of ``supported'' but not, by itself, sufficient.
  \item[Virtual object.] An APFS object whose
    \texttt{obj\_phys\_t} storage flag indicates the object is
    resolvable through an OMAP. \texttt{o\_oid} is independent of
    the block's paddr; the header is validated against
    \texttt{omap\_lookup.oid}.
  \item[\texttt{walk\_skips}.] The fallback walker's per-entry
    skip log. Records \texttt{EACCES}, \texttt{EPERM},
    \texttt{ENOENT}, \texttt{firmlink\_overlay\_source}, and
    \texttt{duplicate\_dir\_inode} reasons with the path and
    keeps the walk going.
  \item[XID.] Transaction identifier. Monotonic per container.
    Used to order checkpoints (chapter~3) and to disambiguate
    OMAP entries (chapter~4).
  \item[Xfield.] The variable-length tail of an FS-tree record
    that carries extended fields. One source-backed cursor rule
    (chapter~6); fail-closed on any malformed shape
    (chapter~7).
\end{description}
