\chapter*{Appendix: Experiment Register}
\addcontentsline{toc}{chapter}{Appendix: Experiment Register}

A reference index of the controlled probes the manual cites. Each
entry names the question the probe answered and the verdict it
produced. Full READMEs, fixture-build scripts, and saved artifacts
live under \path{docs/research/experiments/EX-*/}.

\begin{longtable}{p{0.10\textwidth}p{0.62\textwidth}p{0.20\textwidth}}
\textbf{Probe} & \textbf{Question and result} & \textbf{Verdict} \\
\hline
\endhead

EX-01 &
Does the latest visible NXSB on a mounted lab image stay stable under
concurrent writes? Found that it does not; latest-state raw reads do
not have a stable oracle on live mounted volumes.
& live-state-non-stable \\

EX-02 &
What minimum FS-record taxonomy reproduces the v1 namespace plus
logical-size oracle? Identified DIR\_REC, INODE, dstream fields,
XATTR, SIBLING\_LINK, SIBLING\_MAP as the v1 record families.
& required-record-set \\

EX-03 &
On a detached APFS image, does the raw walker reproduce the mounted
POSIX namespace plus logical size? First positive proof loop.
& pinned-state-parity \\

EX-04 &
Does the EX-03 parity hold across case-insensitive and case-sensitive
APFS variants? Expanded fixture corpus; parity held.
& expanded-corpus-parity \\

EX-05 &
On a still-mounted lab image under write churn, does latest-state
raw output match any baseline or final mounted oracle? Counterexample
result: it does not. Live latest cannot be a v1 model.
& live-latest-not-validatable \\

EX-06 &
Across mutations to a tracked subtree, which identity fields stay
stable? OID stays stable; paddr, XID, checksum, and block hash all
change. Bare-OID cache is unsafe (chapter~10).
& oid-alone-unsafe \\

EX-07 &
Under exact node-identity matches across the full nine-field key,
does subtree-summary reuse ever produce a wrong summary? First run
found zero false reuse in the detached lab corpus. Cache idea
remains alive but unproven for production.
& zero-false-reuse-on-corpus \\

EX-08 &
What does the read-path support matrix look like on a safe host?
Detached unencrypted image-backed APFS is supported for narrow v1.
Mounted images are readable but not supported. Startup-disk raw read
is blocked by privilege.
& read-path-matrix-baseline \\

EX-09 &
Which size metrics belong in v1, and which require their own
oracle? Defined the accounting register: logical size is v1;
allocated, exclusive, shared, snapshot-retained are deferred.
& accounting-register \\

EX-10 &
Does the native Rust checkpoint scanner reach selected NXSB,
checkpoint-map, container OMAP, volume superblock decode, volume
OMAP, and FS-tree root validation on the proof fixture? Yes.
& native-checkpoint-scanner \\

EX-11 &
Does the checkpoint-map ring validate end-to-end with
\texttt{CHECKPOINT\_MAP\_LAST} enforced? Yes, on the proof fixture
and on synthetic malformed cases.
& checkpoint-map-integrity \\

EX-12 &
Does the native OMAP resolver satisfy SR-006 lower-bound
\texttt{(oid, max\_xid)} semantics, with hard stops at the encryption
and unknown-flag cases? Yes; cross-validated against
\texttt{go-apfs identitydump}.
& validated-omap-lookup-contract \\

EX-13 &
Does a Python raw-byte body decoder reconstruct the mounted namespace
and logical-size oracle on the proof fixture? Yes, modulo
candidate-scored xfield layout (later resolved by SR-015 / EX-16).
& validated-native-record-body-contract \\

EX-14 &
Does a second, denser xfield-bearing fixture variant resolve the
xfield layout question? Returned \texttt{oracle\_inconclusive}: Rust
aborted with a checksum mismatch at block 1031 before xfield
comparison could run. Blocker tracked into EX-15.
& oracle-inconclusive \\

EX-15 &
What does block 1031 actually contain, and which of (a)~stale-
checkpoint selection, (b)~missing checkpoint-map rule, (c)~Rust
validation bug, or (d)~malformed source is the cause? Hypothesis
(c). FS-tree internal-node 8-byte values are virtual OIDs requiring
volume-OMAP resolution, not paddrs. Fix landed; two regression tests
added.
& validated-fs-tree-internal-oid-resolution-gap \\

EX-16 &
Does the SR-015 single xfield cursor rule decode every required
xfield in the proof fixture with
\texttt{xf\_used\_data == sum(round\_up(x\_size, 8))} and oracle
parity? Yes, 14 of 14 records.
& validated-sr-015-cursor-rule \\

EX-17 &
Does the Rust body decoder hard-stop on every SR-016 per-record
malformed case with a typed \texttt{InvalidObject}? Yes; 21 unit
tests, all passing.
& validated-sr-016-fail-closed-unit-tests \\

EX-18 &
Does the Rust body decoder produce field-level identical output to
the Python decoder on the proof fixture? Yes: 53 records, identical
\texttt{(node\_paddr, entry\_index)} set, zero divergent fields.
& field-level-parity \\

EX-19 &
Does the SR-017 per-inode logical-size precedence reproduce
\texttt{st\_size} for ordinary, sparse, clone, hard-linked, symlink,
and compressed cases? Yes, 5 of 5 unique inodes; the compressed case
requires \texttt{INODE\_HAS\_UNCOMPRESSED\_SIZE} to beat the decmpfs
header.
& validated-sr-017-precedence \\

EX-20 &
Do Rust paths match POSIX paths byte-for-byte on both case-
insensitive and case-sensitive APFS volumes? Yes; volume case and
normalisation flags propagate correctly. Lookup-by-name still not
claimed.
& validated-sr-018-name-preservation \\

EX-21 &
Does a POSIX-traversal fallback emit the same
\texttt{NamespaceEntry} / \texttt{DirectoryAggregate} shape as Rust
raw mode on the proof fixture? Yes; Python and Rust ports both pass.
& validated-fallback-skeleton \\

EX-22 &
Does the SR-019 per-inode allocated-size precedence reproduce
\texttt{st\_blocks * 512} for ordinary, sparse, clone, hard-link,
symlink, and decmpfs cases? Three of five emit-rows match; sparse
overstates by exactly the sparse-bytes hint; decmpfs is correctly
fail-closed. Sparse case split into a dedicated fail-closed branch
pending an EX-22b sparse corpus.
& partial-validated-sr-019-alloced-size \\

EX-23 &
Does the existing fallback walker produce byte-identical rows when
pointed at a live volume and a read-only snapshot of the same
volume? Probe ran on a clean dev workstation, found only the
sealed-system OS-update snapshot (correctly skipped), and emitted
a blocked verdict with the privileged-rerun reproducer. The shape
claim itself remains unclaimed pending the rerun.
& blocked-no-snapshots-at-all \\

EX-24 &
For the production walker's \texttt{getattrlistbulk} attribute
mask, is the SR-021 vnode-rage hypothesis a measurable lever?
Drec-only and current-mask landed within 1\% on every dimension;
\texttt{fts(3)} was 1.7$\times$ slower than the current mask. The
vnode-rage cost is not load-bearing; the Phase-2 deferred-attribute
refactor is de-prioritised; \texttt{fts} follow-up cancelled.
& vnode-cost-marginal \\

EX-25 &
What is the parallel walker's scaling envelope on Apple silicon at
$T \in \{1, 2, 4, 8, 14\}$? T=4 reaches 2.47$\times$ T=1; T=8
regresses; T=14 catastrophically loses (9.3$\times$ T=1 system CPU
for 1.38$\times$ throughput). The Szorc-2018 / Apple-DTS-2025
container-lock contention regime is real and starts at $T = 2 \times
\text{optimum}$ on this kernel. Production default
\texttt{min(hw.physicalcpu, 4)}.
& validated-parallel-scaling \\

EX-26 &
Against the post-FFI workload (\texttt{/Users/kai/Projects},
320k entries, T=4, warm cache), how much of the post-rewrite
budget is structural rather than kernel-bound, and which
structural rounds clear which fractions of it? Eight rounds
landed against \texttt{TreeNode}, \texttt{NamespaceEntry},
\texttt{Tree::build}, \texttt{build\_aggregates},
\texttt{VisitedDirs}, and the CLI-only aggregates Vec; net
$-32\%$ scan wall, $-50\%$ tree wall, $-28\%$ peak RSS,
allocator pressure roughly halved (chapter~12 §
Structural-Density Pass). EX-26 also closed the SR-019 sparse
+ decmpfs allocated-size precedence: all sparse and decmpfs
shapes match \texttt{st\_blocks * 512} exactly under
\texttt{alloced\_size - sparse\_bytes} and
$\Sigma$ stream-backed-xattr \texttt{alloced\_size}
respectively.
& validated-structural-density-pass; validated-sparse-and-decmpfs \\

EX-27 &
For a fixture with one or more \texttt{cp -c}-created clone
families, can the raw parser walk the extent-reference tree,
recover per-extent refcounts, and compute clone-deduplicated
per-inode bytes that sum to the on-disk allocation? Yes:
\texttt{dstream\_dedup} computed from \texttt{phys\_ext}
records joined with \texttt{file\_extent}s keyed by
\texttt{dstream\_id}; per-inode share is
\texttt{dstream\_dedup / dstream.refcnt}.
$\Sigma$ inode shares equals
$\Sigma$ \texttt{phys\_ext.length} $\cdot$ block size modulo
$\leq$ \texttt{(refcnt - 1)} bytes per clone-shared dstream
(2-byte residue on the fixture). Productive discovery:
\texttt{du -A} on macOS is apparent-size; no public per-file
macOS oracle for clone-dedup exists. The crate ships a
\texttt{real\_size} field on \texttt{NamespaceEntry}.
& validated-clone-dedup \\

EX-28 &
With root privileges, does the existing raw parser produce a
self-consistent shape on a live \texttt{/dev/diskNsM} and
agree with the fallback walker on the same mount? macOS
returned \texttt{EPERM} on the first \texttt{read(2)} of
\texttt{/dev/disk3s1}'s block zero even under \texttt{sudo}.
The kernel's storage-system security policy refuses raw block
access to the live boot data partition independently of POSIX
permissions; SIP and the sealed-system-volume seal both
contribute. Live boot raw mode is not viable on stock Apple
silicon. Contribution: the public
\texttt{apfs\_fastindex::parity} comparator and the env-gated
integration harness, both re-runnable on future hosts.
& live-raw-blocked-by-kernel \\

EX-29 &
What user-visible local snapshots exist on the host, and what
do reclaimable bytes look like? Zero user-visible TM
snapshots; one sealed-system OS-update snapshot
(SR-020-excluded). Reclaimable bytes have no public
read-only macOS oracle (\texttt{tmutil thinlocalsnapshots}
deletes the snapshot it measures), and EX-28's verdict
blocks the raw-extent-diff alternative. Crate ships
\texttt{snapshots::SnapshotVerdict} via
\texttt{tmutil listlocalsnapshots} +
\texttt{diskutil apfs listSnapshots} parsing.
& blocked-no-user-snapshots \\

EX-30 &
Production-shape baseline for the GUI's
\texttt{apfs-fastindex-scan} subprocess on
\texttt{/Users/kai} (the largest single subtree the project
owner routinely scans). Warm median 9.62~s, 162~k entries/s,
1.56~M entries, 172~k aggregates, 392~MB msgpack payload.
The number anchors R4 cache acceptance:
$\leq$1.0~s on unchanged-tree rescan, $\leq$5\% regression on
cold-miss. The R4 v1 cache (\texttt{-{}-cache} flag, ships
this round) delivers 5--50$\times$ on stable trees and
1.0$\times$ on \texttt{/Users/kai}-class churn (the probe
overhead exceeds the walk; mtime churn invalidates every
signature).
& production-baseline-recorded \\

EX-31 &
What's the boot-root semantic the app should target? EX-28
foreclosed the raw-System-vs-raw-Data comparison. The only
deliverable mode is Finder-visible startup root via POSIX
traversal --- which the existing fallback walker already
produces, since POSIX traversal transparently follows
firmlinks. The host's \texttt{st\_dev} reports identically
for every path under \texttt{/} (the kernel hides the
firmlink boundary from \texttt{stat}); the firmlink table at
\texttt{/usr/share/firmlinks} is the authoritative source the
walker already reads at scan-start.
& api-only-already-correct \\

EX-32 &
Research scratchpad: what does the publicly-documented +
reverse-engineered knowledge of APFS encryption cover, and
what would a future feature require? Documentation of the
keybag, per-file \texttt{crypto\_state\_t}, class keys,
\texttt{INCOMPAT\_ENC\_ROLLED}, and the
\texttt{wrapped\_meta\_crypto\_state\_t} of the volume
superblock. Crate exposes \texttt{crypto\_meta} as a
non-production decoder module against synthetic fixtures.
No decryption path is implemented; no feature follows without
an oracle.
& research-scratchpad \\

EX-33 &
Profile the \texttt{apfs\_scan\_search\_names} FFI on the
1.56~M-entry \texttt{/Users/kai} baseline; ship one lift if
there's headroom. The pre-lift \texttt{Vec<String>} +
\texttt{str::contains} ran 30--55~ms median per keystroke;
the lifted flat-\texttt{Vec<u8>}-buffer +
\texttt{memchr::memmem::Finder} runs 9--12~ms median for
typical 4--12-character queries, $>$12$\times$ on zero-match
queries (memory-bandwidth floor). Comfortably inside the
``feels live'' threshold.
& validated-flat-buffer-memchr-lift \\

\end{longtable}
